---
title: mlr3 Basics on "Iris" - Hello World!
slug: mlr3-basics-iris
categories:
  - mlr3
  - basics
  - iris data set
  - classification
description: |
  Basic ML operations on iris: Train, predict, score, resample and benchmark. A simple, hands-on intro to mlr3.
author:
  - name: Bernd Bischl
date: 03-18-2020
output:
  distill::distill_article:
    toc: true
    self_contained: false
    highlight_downlit: true
    css: ../../custom.css
---

```{r 2020-03-18-iris-mlr3-basics-001, include=FALSE}
 knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 120)
)

library(mlr3book)
library(kableExtra)

set.seed(7832)

ggplot2::theme_set(ggplot2::theme_light())
```

## Goals and Prerequisites

This use case shows how to use the basic `r mlr_pkg("mlr3")` package on the iris `r ref("Task")`, so it's our "Hello World" example.
It assumes no prior knowledge in ML or `r mlr_pkg("mlr3")`.
You can find most of the content here also in the [mlr3book](https://mlr3book.mlr-org.com/) in a more detailed way.
Hence we will not make a lot of general comments, but keep it hands-on and short.

The following operations are shown:

* Creating `r ref("Task", "Tasks")` and `r ref("Learner", "Learners")`
* Training and predicting
* `r ref("Resampling")` / `r ref("ResamplingCV", "cross-validation")`
* Installing more `r ref("Learner", "Learners")`
* `r ref("benchmark", "Benchmarking")` to compare multiple `Learners`

## Loading basic packages

We load the `r mlr_pkg("mlr3verse")` package which pulls in the most important packages for this example.
The `r mlr_pkg("mlr3learners")` package loads additional `r ref("Learner", "learners")`.

```{r 2020-03-18-iris-mlr3-basics-002}
library(mlr3verse)
library(mlr3learners)
```

## Creating tasks and learners

Let's work on the canonical, simple iris data set, and try out some ML algorithms.
We will start by using a decision tree with default settings.

```{r 2020-03-18-iris-mlr3-basics-003}
# creates mlr3 task from scratch, from a data.frame
# 'target' names the column in the dataset we want to learn to predict
task = as_task_classif(iris, target = "Species")
# in this case we could also take the iris example from mlr3's dictionary of shipped example tasks
# 2 equivalent calls to create a task. The second is just sugar for the user.
task = mlr_tasks$get("iris")
task = tsk("iris")
print(task)
# create learner from dictionary of mlr3learners
# 2 equivalent calls:
learner_1 = mlr_learners$get("classif.rpart")
learner_1 = lrn("classif.rpart")
print(learner_1)
```

## Train and predict

Now the usual ML operations: Train on some observations, predict on others.

```{r 2020-03-18-iris-mlr3-basics-004}
# train learner on subset of task
learner_1$train(task, row_ids = 1:120)
# this is what the decision tree looks like
print(learner_1$model)
# predict using observations from task
prediction = learner_1$predict(task, row_ids = 121:150)
# predict using "new" observations from an external data.frame
prediction = learner_1$predict_newdata(newdata = iris[121:150, ])
print(prediction)
```

## Evaluation

Let's score our `r ref("Prediction")` object with some metrics.
And take a deeper look by inspecting the confusion matrix.

```{r 2020-03-18-iris-mlr3-basics-005}
head(as.data.table(mlr_measures))
scores = prediction$score(msr("classif.acc"))
print(scores)
scores = prediction$score(msrs(c("classif.acc", "classif.ce")))
print(scores)
cm = prediction$confusion
print(cm)
```

## Changing hyperpars

The `r ref("Learner")` contains information about all parameters that can be configured, including data type, constraints, defaults, etc.
We can change the hyperparameters either during construction of later through an active binding.

```{r 2020-03-18-iris-mlr3-basics-006, eval = FALSE}
as.data.table(learner_1$param_set)
```

```{r 2020-03-18-iris-mlr3-basics-007, echo = FALSE}
as.data.table(learner_1$param_set)[, .(id, class, lower, upper, nlevels)] %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE)
```

```{r 2020-03-18-iris-mlr3-basics-008}
learner_2 = lrn("classif.rpart", predict_type = "prob", minsplit = 50)
learner_2$param_set$values$minsplit = 50
```

## Resampling

`r ref("Resampling")` simply repeats the train-predict-score loop and collects all results in a nice `r ref("data.table::data.table()")`.

```{r 2020-03-18-iris-mlr3-basics-009}
cv10 = rsmp("cv", folds = 10)
rr = resample(task, learner_1, cv10)
print(rr)
```

```{r 2020-03-18-iris-mlr3-basics-010, eval = FALSE}
rr$score(msrs(c("classif.acc", "classif.ce")))
```

```{r 2020-03-18-iris-mlr3-basics-011, echo = FALSE}
rr$score()[, .(iteration, task_id, learner_id, resampling_id, classif.ce)] %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE)
```

```{r 2020-03-18-iris-mlr3-basics-012, results = 'hide'}
# get all predictions nicely concatenated in a table
prediction = rr$prediction()
head(as.data.table(prediction))
```

```{r 2020-03-18-iris-mlr3-basics-013, echo = FALSE}
head(as.data.table(prediction)) %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE)
```

```{r 2020-03-18-iris-mlr3-basics-014}
cm = prediction$confusion
print(cm)
```

## Populating the learner dictionary

`r mlr_pkg("mlr3learners")` ships out with a dozen different popular `r ref("Learner", "Learners")`.
We can list them from the dictionary.
If we want more, we can install an extension package, `r mlr_pkg("mlr3extralearners")`, from GitHub.
Note how after the installation the dictionary increases in size.

```{r 2020-03-18-iris-mlr3-basics-015, message = FALSE}
head(as.data.table(mlr_learners)[, c("key", "packages")])
# remotes::install_github("mlr-org/mlr3extralearners")
library(mlr3extralearners)
print(as.data.table(mlr_learners)[, c("key", "packages")])
```

## Benchmarking multiple learners

The `r ref("benchmark")` function can conveniently compare `r ref("Learner", "Learners") on the same dataset(s).

```{r 2020-03-18-iris-mlr3-basics-016}
learners = list(learner_1, learner_2, lrn("classif.randomForest"))
grid = benchmark_grid(task, learners, cv10)
bmr = benchmark(grid)
print(bmr)
print(bmr$aggregate(measures = msrs(c("classif.acc", "classif.ce"))))
```

## Conclusion

We left out a lot of details and other features.
If you want to know more, read the [mlr3book](https://mlr3book.mlr-org.com/) and the documentation of the mentioned packages.
