---
title: Imbalanced Data Handling with mlr3
categories:
  - classification
  - imbalanced data
  - mlr3tuning
  - tuning
  - optimization
  - nested resampling
  - optdigits data set
  - classification
author:
  - name: Giuseppe Casalicchio
date: 03-30-2020
description: |
  This use case compares different approaches to handle class imbalance for the optdigits binary classification data set.
output:
  distill::distill_article:
    toc: true
    self_contained: false
    highlight_downlit: true
    css: ../../custom.css
bibliography: bibliography.bib
params:
  eval_all: FALSE
---

```{r 2020-03-30-imbalanced-data-001, include=FALSE}
 knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 120)
)

library(mlr3book)
library(kableExtra)

set.seed(7832)

ggplot2::theme_set(ggplot2::theme_light())
```

```{r 2020-03-30-imbalanced-data-002, include=FALSE, eval = params$eval_all}
future::plan(list("sequential", "multisession"))
```

# Intro

This use case compares different approaches to handle class imbalance for the
`r ref("mlr_tasks_optdigits", "Optical Recognition of Handwritten Digits")` (optdigits) binary classification data set using the `r mlr_pkg("mlr3")` package.
We mainly focus on undersampling the majority class, oversampling the minority class, and the SMOTE imbalance correction [@smote] that enriches the minority class with synthetic data.
The use case requires prior knowledge in basic ML concepts (issues imbalanced data, hyperparameter tuning, nested cross-validation).
The R packages `r mlr_pkg("mlr3")`, `r mlr_pkg("mlr3pipelines")` and `r mlr_pkg("mlr3tuning")` will be used.
You can find most of the content here also in the [mlr3book](https://mlr3book.mlr-org.com/) explained in a more detailed way.

These steps are performed:

* Retrieve data sets from `OpenML`
* Define imbalance correction pipeline `r ref("Graph")`s (undersampling, oversampling and SMOTE) with `r mlr_pkg("mlr3pipelines")`
* Autotune the `r ref("Graph")` together with a learner using `r mlr_pkg("mlr3tuning")`
* Benchmark the autotuned `r ref("Graph")` and visualize the results using `r mlr_pkg("mlr3viz")`

# Prerequisites

We load the most important packages for this post.

```{r 2020-03-30-imbalanced-data-003}
library(mlr3verse)
library(mlr3learners)
library(OpenML)
```

We initialize the random number generator with a fixed seed for reproducibility, and decrease the verbosity of the logger to keep the output clearly represented.

```{r 2020-03-30-imbalanced-data-004}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
```

# Retrieve data sets from OpenML

[OpenML.org](https://www.openml.org) is an open machine learning platform, which allows users to share data, code and machine learning experiments.
The OpenML R package can query available data sets using a filter-like approach by providing desired dataset characteristics like `number.of.classes` or `number.of.features`.

```{r 2020-03-30-imbalanced-data-005, eval = FALSE}
# get list of curated binary classification data sets (see https://arxiv.org/abs/1708.03731v2)
ds = listOMLDataSets(
  number.of.classes = 2,
  number.of.features = c(1, 100),
  number.of.instances = c(5000, 10000)
)
# select imbalanced data sets (without categorical features as SMOTE cannot handle them)
ds = subset(ds, minority.class.size / number.of.instances < 0.2 &
  number.of.symbolic.features == 1)
ds

# pick one data set from list above
d = getOMLDataSet(980)
d
```

After downloading the chosen data set, we create an `r mlr_pkg("mlr3")` classification task:

```{r 2020-03-30-imbalanced-data-006, eval = FALSE}
# make sure target is a factor and create mlr3 tasks
data = as.data.frame(d)
data[[d$target.features]] = as.factor(data[[d$target.features]])
task = as_task_classif(data, id = d$desc$name, target = d$target.features)
task
```

Please note that the optdigits dataset is also included in the `r mlr_pkg("mlr3data")`  package where you can get the preprocessed (integers properly encoded as such, etc.) data via:

```{r 2020-03-30-imbalanced-data-007}
data("optdigits", package = "mlr3data")
task = as_task_classif(optdigits, target = "binaryclass", positive = "P")
```

Quick overview of the data:

```{r 2020-03-30-imbalanced-data-008}
skimr::skim(task$data())
```

# Imbalance correction

In `r mlr_pkg("mlr3pipelines")` , there is a `r ref("PipeOpClassBalancing", "class balancing")` and a `r ref("PipeOpSmote", "smote")` pipe operator that can be combined with any learner.
Below, we define the undersampling, oversampling and SMOTE `r ref("PipeOp")`s/`r ref("Graph")`.
All three imbalance correction methods have hyperparameters to control the degree of class imbalance.
We apply the `r ref("PipeOp")`s/`r ref("Graph")` to the current `r ref("Task", "task")` with specific hyperparameter values to see how the class balance changes:

```{r 2020-03-30-imbalanced-data-009}
# check original class balance
table(task$truth())

# undersample majority class (relative to majority class)
po_under = po("classbalancing",
  id = "undersample", adjust = "major",
  reference = "major", shuffle = FALSE, ratio = 1 / 6)
# reduce majority class by factor '1/ratio'
table(po_under$train(list(task))$output$truth())

# oversample majority class (relative to majority class)
po_over = po("classbalancing",
  id = "oversample", adjust = "minor",
  reference = "minor", shuffle = FALSE, ratio = 6)
# enrich minority class by factor 'ratio'
table(po_over$train(list(task))$output$truth())
```

Note that the original SMOTE algorithm only accepts numeric features (see `r ref("smotefamily::SMOTE")`).
To keep it simple, we therefore preprocess the data and coerce integers to numerics prior to running `SMOTE` and reverse this afterwards:

```{r 2020-03-30-imbalanced-data-010}
# SMOTE enriches the minority class with synthetic data
gr_smote =
  po("colapply", id = "int_to_num",
    applicator = as.numeric, affect_columns = selector_type("integer")) %>>%
  po("smote", dup_size = 6) %>>%
  po("colapply", id = "num_to_int",
    applicator = function(x) as.integer(round(x, 0L)), affect_columns = selector_type("numeric"))
# enrich minority class by factor (dup_size + 1)
table(gr_smote$train(task)[[1L]]$truth())
```

# Construct `AutoTuner`

We combine the `r ref("PipeOp")`s/`r ref("Graph")` with a learner (here `r ref("mlr_learners_classif.ranger", "ranger")`) to make each pipeline graph behave like a learner:

```{r 2020-03-30-imbalanced-data-011}
# create random forest learner
learner = lrn("classif.ranger", num.trees = 50)

# combine learner with pipeline graph
learner_under = as_learner(po_under %>>% learner)
learner_under$id = "undersample.ranger"
learner_over = as_learner(po_over %>>% learner)
learner_over$id = "oversample.ranger"
learner_smote = as_learner(gr_smote %>>% learner)
learner_smote$id = "smote.ranger"
```

We define the search space in order to tune the hyperparameters of the class imbalance methods.

```{r 2020-03-30-imbalanced-data-012}
# define parameter search space for each method
search_space_under = ps(undersample.ratio = p_dbl(1 / 6, 1))
search_space_over = ps(oversample.ratio = p_dbl(1, 6))
search_space_smote = ps(
  smote.dup_size = p_int(1, 6),
  smote.K = p_int(1, 6),
  # makes sure we use numbers to the power of two to better explore the parameter space
  .extra_trafo = function(x, param_set) {
    x$smote.K = round(2^(x$smote.K))
    x
  }
)
```

We create an `r ref("AutoTuner")` class from the learner to tune the graph (random forest learner + imbalance correction method) based on a 3-fold cross-validation using the `r ref("mlr_measures_classif.fbeta", "F-beta Score")` as performance measure.
To keep runtime low, we define the search space only for the imbalance correction method.
However, one can also jointly tune the hyperparameter of the learner along with the imbalance correction method by extending the search space with the learner's hyperparameters.
Note that SMOTE has two hyperparameters `K` and `dup_size`.
While `K` changes the behavior of the SMOTE algorithm, `dup_size` will affect oversampling rate.
To focus on the effect of the oversampling rate on the performance, we will consider SMOTE with K = 2 as a different imbalance correction method as SMOTE with K = 4 (and so on).
Hence, we use grid search with 5 different hyperparameter configurations for the undersampling method, the oversampling method and each SMOTE variant for tuning:

```{r 2020-03-30-imbalanced-data-013}
inner_cv3 = rsmp("cv", folds = 3)
measure = msr("classif.fbeta")

learns = list(
  auto_tuner(
    method = "grid_search",
    learner = learner_under,
    resampling = inner_cv3,
    measure = measure,
    search_space = search_space_under,
    resolution = 6
  ),
  auto_tuner(
    method = "grid_search",
    learner = learner_over,
    resampling = inner_cv3,
    measure = measure,
    search_space = search_space_over,
    resolution = 6
  ),
  auto_tuner(
    method = "grid_search",
    learner = learner_smote,
    resampling = inner_cv3,
    measure = measure,
    search_space = search_space_smote,
    resolution = 6
  )
)
```

# Benchmark `AutoTuner`

The `r ref("AutoTuner")` is a fully tuned graph that behaves like a usual learner.
For the tuning a 3-fold CV is used.
Now, we use the `r ref("benchmark()")` function to compare the tuned class imbalance pipeline graphs based on a holdout for the outer evaluation:

```{r 2020-03-30-imbalanced-data-014}
outer_holdout = rsmp("holdout")
design = benchmark_grid(
  tasks = task,
  learners = learns,
  resamplings = outer_holdout
)
print(design)
```

```{r 2020-03-30-imbalanced-data-015, eval = params$eval_all}
# NOTE: This code runs about 5 minutes
bmr = benchmark(design, store_models = TRUE)
```

```{r 2020-03-30-imbalanced-data-016, echo = FALSE, eval = params$eval_all}
saveRDS(bmr, "data/bmr.rda")
```

```{r 2020-03-30-imbalanced-data-017, echo = FALSE}
bmr = readRDS("data/bmr.rda")
```

## Visualize benchmark results

```{r 2020-03-30-imbalanced-data-018}
bmr$aggregate(measure)
# one value per boxplot since we used holdout as outer resampling
autoplot(bmr, measure = measure)
```

## Visualize the tuning path

With `store_models = TRUE` we allow the `r ref("benchmark()")` function to store each single model that was computed during tuning.
Therefore, we can plot the tuning path of the best learner from the subsampling iterations:

```{r 2020-03-30-imbalanced-data-019}
library(ggplot2)
bmr_data_learners = as.data.table(bmr)$learner
utune_path = bmr_data_learners[[1]]$model$tuning_instance$archive$data
utune_gg = ggplot(utune_path, aes(x = undersample.ratio, y = classif.fbeta)) +
  geom_point(size = 3) +
  geom_line() + ylim(0.9, 1)

otune_path = bmr_data_learners[[2]]$model$tuning_instance$archive$data
otune_gg = ggplot(otune_path, aes(x = oversample.ratio, y = classif.fbeta)) +
  geom_point(size = 3) +
  geom_line() + ylim(0.9, 1)

stune_path = bmr_data_learners[[3]]$model$tuning_instance$archive$data
stune_gg = ggplot(stune_path, aes(
  x = smote.dup_size,
  y = classif.fbeta, col = factor(smote.K))) +
  geom_point(size = 3) +
  geom_line() + ylim(0.9, 1)

library(ggpubr)
ggarrange(utune_gg, otune_gg, stune_gg, common.legend = TRUE, nrow = 1)
```

The results show that oversampling the minority class (for simple oversampling as well as for SMOTE) and undersampling the majority class yield a better performance for this specific data set.

# Conclusion

In this post, we tuned and compared 5 different settings of sampling ratios for the undersampling method, the oversampling method and different SMOTE variants (using different values of `K` nearest neighbors during the sampling process).
If you want to know more, read the [mlr3book](https://mlr3book.mlr-org.com/) and the documentation of the mentioned packages.
