---
title: A pipeline for the titanic data set - Imputation 
categories:
  - imputation
  - classification
  - mlr3pipelines
  - feature engineering
author:
  - name: Florian Pfisterer
description: |
  This post shows how to build a Graph using the mlr3pipelines package on the "titanic" dataset.
date: 04-27-2020
output:
  distill::distill_article:
    self_contained: false
    css: ../../custom.css
---

This is the second post of the titanic use case series.
You can find the first use case [here](/2020-03-12-intro-pipelines-titanic/).
<!-- Need to fix hyperlink -->
In this section we will focus on the usage of `mlr3pipelines`.
Specifically, this section illustrates the different options when it comes to data imputation.
Furthermore, the section on benchmarking shows how to visualize our and compare our results.
Again, we need to take some prepatory steps: 

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Just some preparation
knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 80)
)
```

For the case that you are only tuning in for the second part, we need to load some packages: 
The data is part of the `mlr3data` package.

```{r}
library("mlr3")
library("mlr3learners")
library("mlr3pipelines")
library("mlr3data")
library("mlr3misc")
library("ggplot2")
library("mlr3viz")
data("titanic")
```

## Imputation

A very simple way to do this to just impute a constant value for each feature.
We could i.e. impute every `character` or `factor` column  with `missing` and every numeric column with `-999`.
And depending on the model, this might actually be fine.
This approach has a few drawbacks though:

* `-999` could be a real value in the data.
* imputing `-999` skews the distribution of the data, which might result in bad models.

As a result, instead of imputing a constant value, we will do two things:
* Draw samples from each numeric features' histogram using `PipeOpImputeHist`
* Add an additional column for each `variable` that indicates whether a value was missing or not.
  If the information that a value was missing is important, this column contains this information.

This imputation scheme is called 'imputation with constants' and is already implemented in `mlr3pipelines`.
It can be done using `PipeOpImputeConstant`.

Before imputation, our data looks as follows:

```{r}
task$missings()
```


Let's first deal with the categorical variables:

```{r}
po_newlvl <- po("imputenewlvl")
task_newlvl <- po_newlvl$train(list(task))[[1]]
```

Note that we use the `PipeOp` in an unusual way, which is why the syntax does not look very clean.
We'll learn how to use a full graph below.

First, let's look at the result:

```{r}
task_newlvl$missings()
```

Cool! `embarked` does not have missing values anymore. 
Note that `PipeOpImputeNewLvl` by default affects `character`, `factor` and `ordered` columns.

For the `numeric` features we want to do two things, impute values and add an indicator column.
In order to do this, we need a more complicated structure, a `Graph`.

Our `po_indicator` creates the indicator column.
We tell it to only do this for `numeric` and `integer` columns via its `param_vals`, and additionally tell it to create a numeric column (0 = "not missing", 1 = "missing").

```{r}
po_indicator <- po("missind",
  param_vals = list(affect_columns = selector_type(c("numeric", "integer")), type = "numeric")
)
```

Now we can simultaneously impute features from the histogram and create indicator columns.
This can be achieved using the `gunion` function, which puts two operations in parallel:

```{r}
gr <- gunion(list(po_indicator, po("imputehist")))
gr <- gr %>>% po("featureunion")
```

Afterwards, we `cbind` the resulting data using `po("featureunion")`, connecting the different operations using our **graph connector**: `%>>%`.
We can now also connect the newlvl imputation:

```{r}
gr <- gr %>>% po("imputenewlvl")
```

and see what happens when we now train the whole **Graph**:

```{r}
task_imputed <- gr$clone()$train(task)[[1]]
task_imputed$missings()
```

Awesome, now we do not have any missing values!

```{r}
autoplot(task_imputed)
```

We could now use `task_imputed` for resampling and see whether a **ranger** model does better.
But this is dangerous! 
If we preprocess all training data at once, data could leak through the different cross-validation folds.
In order to do this properly, we have to process the training data in every fold separately.
Luckily, this is automatically handled in our `Graph`, if we use it through a `GraphLearner`.

We can simple append a `ranger` learner to the Graph and create a `GraphLearner` from this.

```{r}
glrn <- GraphLearner$new(gr$clone() %>>% po(learner))
```

```{r}
res <- resample(task, glrn, rdesc, store_models = TRUE)
agg2 <- res$aggregate(msr("classif.acc"))
agg2
```

So our model has not improved heavily, currently it has an accuracy of  `r round(agg, 3)`.
In order to improve more, we might need to do some feature engineering.

## Feature Engineering

We will do this using `PipeOpMutate` in order to showcase the power of `mlr3pipelines`.
Additionally, we will make use of the `character` columns.
Hence, we will re-select them:

```{r}
task$col_roles$feature <- c(task$feature_names, c("cabin", "name", "ticket"))
```


```{r}
library("stringi")
po_ftextract <- po("mutate", param_vals = list(
  mutation = list(
    fare_per_person = ~ fare / (parch + sib_sp + 1),
    deck = ~ factor(stri_sub(cabin, 1, 1)),
    title = ~ factor(stri_match(name, regex = ", (.*)\\.")[, 2]),
    surname = ~ factor(stri_match(name, regex = "(.*),")[, 2]),
    ticket_prefix = ~ factor(stri_replace_all_fixed(stri_trim(stri_match(ticket, regex = "(.*) ")[, 2]), ".", ""))
  )
))
```

Quickly checking what happens:

```{r}
task_eng <- po_ftextract$clone()$train(list(task))[[1]]
task_eng$data()
```

```{r}
autoplot(task_eng$clone()$select(c("sex", "age")), type = "pairs")
```

Now we can put everything together again, we concatenate our new `PipeOp` with the `Graph` created above and use `PipeOpSelect` in order to de-select the `character` features we used for feature extraction.
Additionally, we collapse the 'surname', so only surnames that make up more than 0.6 \% of the data are kept.

In summary we do the following: 

* `mutate`: The `po_ftextract` we defined above extracts additional features from the data.
* `collapsefactors`: Removes factor levels that make up less then 3 \% of the data.
* `select`: Drops `character` columns.
* `gunion`: Puts two `PipeOp`s in parallel.
  * `missind`: `po_indicator` adds a column for each numeric with the info whether the value is NA or not.
  * `imputehist`: Imputes numeric and integer columns by sampling from the histogram.
* `featureunion`: Cbind's parallel data streams.
* `imputenewlvl`: Imputes factor and ordered columns.
* `fixfactors`: Removes empty factor levels and removes factor levels that do not exist during training.
* `imputesample`: In some cases, if missing factor levels do not occur during training but only while predicting, `imputenewlvl` does not create a new level. For those, we sample a random value.
* `learner`: Appends a learner to the `Graph`.

The full graph we created is the following:

```{r}
learner <- lrn("classif.ranger")
learner$param_set$values <- list(num.trees = 500, min.node.size = 4)
```

```{r}
gr_final <- po_ftextract %>>%
  po("collapsefactors", param_vals = list(no_collapse_above_prevalence = 0.03)) %>>%
  po("select", param_vals = list(selector = selector_invert(selector_type("character")))) %>>%
  gunion(list(po_indicator, po("imputehist"))) %>>%
  po("featureunion") %>>%
  po("imputenewlvl") %>>%
  po("fixfactors") %>>%
  po("imputesample") %>>%
  po(learner)
```

## Evaluation 

Let us see if things have improved:

```{r}
glrn <- GraphLearner$new(gr_final)
res <- resample(task, glrn, rdesc, store_models = TRUE)
agg_ranger <- res$aggregate(msr("classif.acc"))
agg_ranger
```

We have improved even more!

## Benchmarking 

To undertake benchmarking, we need to set up a benchmarking design.
The first step is creating a list with the learners we used, namely the learners form the first and second part of this use case.

```{r}
lrn_list <- list(
  lrn("classif.rpart", predict_type = "prob"),
  lrn("classif.ranger", predict_type = "prob")
)
```

After having done that, we can set up our benchmark grid: 
```{r}
bm_design <- benchmark_grid(task = task_imputed, resamplings = rsmp("cv", folds = 10), learners = lrn_list)
bmr <- benchmark(bm_design, store_models = TRUE)
print(bmr)
```

So, where do we go from here?
We could for instance use a boxplot and compare the two: 

```{r}
autoplot(bmr) + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Further we are able to compare sensitivity and specificity: 

```{r}
autoplot(bmr$clone()$filter(task_id = "titanic"), type = "roc")
```

Moreover, one can compare the the precision-recall: 

```{r}
# Precision vs Recall
ggplot2::autoplot(bmr, type = "prc")
```

As one can see, there are various options when it comes to benchmarking and visualizing.
You could have a look at some other use cases in our galary for inspiration.

## Future

In this case we have examined a number of different features, but there are many more things to explore!
We could extract even more information from the different features and see what happens.
But now you are left to yourself! There are many [kaggle kernels](https://www.kaggle.com/c/titanic) that treat the **Titanic Dataset**
available. This can be a great starter to find even better models.
