---
title: Integer Hyperparameters in Tuners for Real-valued Search Spaces
categories:
  - mlr3tuning
  - tuning
  - optimization
  - iris data set
  - classification
author:
  - name: Marc Becker
description: |
  How to tune integer hyperparameters with tuners that can only propose real numbers.
date: 01-19-2021
output:
  distill::distill_article:
    toc: true
    self_contained: false
    highlight_downlit: true
    css: ../../custom.css
---

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-001, include=FALSE}
 knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 120)
)

library(mlr3book)
library(kableExtra)

ggplot2::theme_set(ggplot2::theme_light())
```

## Introduction

`r ref("Tuner")` for real-valued search spaces are not able to tune on integer hyperparameters.
However, it is possible to round the real values proposed by a `r ref("Tuner")` to integers before passing them to the learner in the evaluation.
We show how to apply a parameter transformation to a `r ref("ParamSet")` and use this set in the tuning process.

We load the `r mlr_pkg("mlr3verse")` package which pulls in the most important packages for this example.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-002}
library(mlr3verse)
```

We initialize the random number generator with a fixed seed for reproducibility, and decrease the verbosity of the logger to keep the output clearly represented.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-003}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

## Task and Learner

In this example, we use the k-Nearest-Neighbor classification learner.
We want to tune the integer-valued hyperparameter `k` which defines the numbers of neighbors.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-004}
learner = lrn("classif.kknn")
print(learner$param_set$params$k)
```

# Tuning

We choose generalized simulated annealing as tuning strategy.
The `param_classes` field of `r ref("TunerGenSA")` states that the tuner only supports real-valued (`ParamDbl`) hyperparameter tuning.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-005}
print(tnr("gensa"))
```

To get integer-valued hyperparameter values for `k`, we construct a search space with a transformation function.
The `as.integer()` function converts any real valued number to an integer by removing the decimal places.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-006}
search_space = ps(
  k = p_dbl(lower = 3, upper = 7.99, trafo = as.integer)
)
```

We start the tuning and compare the results of the search space to the results in the space of the learners hyperparameter set.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-007}
instance = tune(
  method = "gensa",
  task = tsk("iris"),
  learner = learner,
  resampling = rsmp("holdout"),
  measure = msr("classif.ce"),
  term_evals = 20,
  search_space = search_space)
```

The optimal `k` is still a real number in the search space.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-008}
instance$result_x_search_space
```

However, in the learners hyperparameters space, `k` is an integer value.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-009}
instance$result_x_domain
```

The archive shows us that for all real-valued `k` proposed by `r cran_pkg("GenSA")`, an integer-valued `k` in the learner hyperparameter space (`x_domain_k`) was created.

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-010, eval = FALSE}
as.data.table(instance$archive)
```

```{r integer-hyperparameters-in-tuners-for-real-valued-search-spaces-011, echo = FALSE}
as.data.table(instance$archive)[, .(k, classif.ce, x_domain_k)] %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE)
```

Internally, `r ref("TunerGenSA")` was given the parameter types of the search space and therefore suggested real numbers for `k`.
Before the performance of the different `k` values was evaluated, the transformation function of the `search_space` parameter set was called and `k` was transformed to an integer value.

Note that the tuner is not aware of the transformation.
This has two problematic consequences:
First, the tuner might propose different real valued configurations that after rounding end up to be already evaluated configurations and we end up with re-evaluating the same hyperparameter configuration.
This is only problematic, if we only optimze integer parameters.
Second, the rounding introduces discontinuities which can be problematic for some tuners.

We successfully tuned a integer-valued hyperparameter with `r ref("TunerGenSA")` which is only suitable for an real-valued search space.
This technique is not limited to tuning problems.
`r ref("Optimizer")` in `r mlr_pkg("bbotk")` can be also used in the same way to produce points with integer parameters.
