---
title: A pipeline for the titanic data set
categories:
  - imputation
  - classification
  - mlr3pipelines
  - feature engineering
author:
  - name: Florian Pfisterer
description: |
  This post shows how to build a Graph using the mlr3pipelines package on the "titanic" dataset.
date: 03-12-2020
output:
  distill::distill_article:
    self_contained: false
    css: ../../custom.css
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Just some preparation
knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 80)
)
if (require("data.table")) data.table::setDTthreads(1)
options(width = 90)
set.seed(8008135)
lgr::get_logger("mlr3")$set_threshold("warn")
library(skimr)
```

## Intro

First of all we are going to load required packages and the data. 
The data is part of the `mlr3data` package.

```{r}
library("mlr3")
library("mlr3learners")
library("mlr3pipelines")
library("mlr3data")
library("mlr3misc")
data("titanic")
```

The titanic data is very interesting to analyze, even though it is part of many tutorials 
and showcases. 
This is because it requires many steps often required in real-world applications of machine 
learning techniques, such as **feature engineering**, **missing value imputation**, **handling factors**
and others.


In order to obtain solutions comparable to official leaderboards, such as the ones available from CRAN, we split the data into train and test set before doing any further analysis.

```{r}
titanic_train = titanic[1:891, ]
titanic_test = titanic[892:1309, ]
```

## Exploratory Data Analysis

With the dataset, we get an explanation of the meanings of the different variables:

```
survived        Survival
                (0 = No; 1 = Yes)
pclass          Passenger Class
                (1 = 1st; 2 = 2nd; 3 = 3rd)
name            Name
sex             Sex
age             Age
sibsp           Number of Siblings/Spouses Aboard
parch           Number of Parents/Children Aboard
ticket          Ticket Number
fare            Passenger Fare
cabin           Cabin
embarked        Port of Embarkation
                (C = Cherbourg; Q = Queenstown; S = Southampton)
```

We can use the `skimr` package in order to get a first overview of the data:

```{r, R.options=list(width = 120)}
skimr::skim(titanic_train)
skimr::skim(titanic_test)
```

Here we can also inspect the data for differences in the train and test set. 
This might be important, as shifts in the data distribution often make our models unreliable.

```{r, out.width="100%", fig.height=7}
DataExplorer::plot_bar(titanic_train, nrow = 5, ncol = 3)
```

```{r, out.width="100%", fig.height=4, warning=FALSE}
DataExplorer::plot_histogram(titanic_train, nrow = 2, ncol = 3)
DataExplorer::plot_boxplot(titanic_train, by = "survived", nrow = 2, ncol = 3)
```

We can now create a `Task` from our data. 
As we want to classify whether the person survived or not, we will create a 
`TaskClassif`. We'll ignore the 'titanic_test' data for now and come back to it later.

## A first model

```{r}
task = TaskClassif$new("titanic", titanic_train, target = "survived", positive = "1")
task
```

Our `Task` currently has $3$ features of type `character`, which we don't really know how  to handle:
"Cabin", "Name", "Ticket" and "PassengerId".
Additionally, from our `skim` of the data, we have seen, that they have many unique values (up to 891).

We'll drop them for now and see how we can deal with them later on.

```{r}
task$select(cols = setdiff(task$feature_names, c("cabin", "name", "ticket")))
```

Additionally, we create a resampling instance that allows to compare data.

```{r}
rdesc = rsmp("cv", folds = 3L)$instantiate(task)
```

To get a first impression of what performance we can fit a simple decision tree:

```{r}
learner = mlr_learners$get("classif.rpart")
# or shorter:
learner = lrn("classif.rpart")

res = resample(task, learner, rdesc, store_models = TRUE)
agg = res$aggregate(msr("classif.acc"))
agg
```

So our model should have a minimal accuracy of `r round(agg, 3)` in order to improve over the
simple decision tree.

If we now try to fit a 'ranger' random forest model, we will get an error, 
as 'ranger' models can not naturally handle missing values.

```{r, error = TRUE}
learner = lrn("classif.ranger")
learner$param_set$values = list(num.trees = 250, min.node.size = 4)
res = resample(task, learner, rdesc, store_models = TRUE)
```

This means we have to find a way to impute the missing values.

## Imputation

A very simple way to do this to just impute a constant value for each 
feature, we could i.e. impute every `character` or `factor` column  with `missing` and 
every numeric column with `-999`.
And depending on the model, this might actually be fine.
This approach has a few drawbacks though:

* `-999` could be a real value in the data.
* imputing `-999` skews the distribution of the data, which might result in bad models.

As a result, instead of imputing a constant value, we will do two things:
* Draw samples from each numeric features' histogram using `PipeOpImputeHist`
* Add an additional column for each `variable` that indicates whether a value was missing or not.
  If the information that a value was missing is important, this column contains this information.

This imputation scheme is called 'imputation with constants' and is already implemented in `mlr3pipelines`.
It can be done using `PipeOpImputeConstant`.

Before imputation, our data looks as follows:

```{r}
task$missings()
```

Let's first deal with the categorical variables:

```{r}
po_newlvl = po("imputenewlvl")
task_newlvl = po_newlvl$train(list(task))[[1]]
```

Note that we use the `PipeOp` in an unusual way, which is why the syntax does not look
very clean. We'll learn how to use a full graph below.

First, let's look at the result:

```{r}
task_newlvl$missings()
```

Cool! `embarked` does not have missing values anymore. Note that `PipeOpImputeNewLvl` by default
affects `character`, `factor` and `ordered` columns.

For the `numeric` features we want to do two things, impute values and add an indicator column.
In order to do this, we need a more complicated structure, a `Graph`.

Our `po_indicator` creates the indicator column. We tell it to only do this for `numeric` and `integer` columns
via its `param_vals`, and additionally tell it to create a numeric column (0 = "not missing", 1 = "missing").

```{r}
po_indicator = po("missind",
  param_vals = list(affect_columns = selector_type(c("numeric", "integer")), type = "numeric"))
```

Now we can simultaneously impute features from the histogram and create indicator columns.
This can be achieved using the `gunion` function, which puts two operations in parallel:

```{r}
gr = gunion(list(po_indicator, po("imputehist")))
gr = gr %>>% po("featureunion")
```

Afterwards, we `cbind` the resulting data using `po("featureunion")`, connecting the different 
operations using our **graph connector**: `%>>%`.

We can now also connect the newlvl imputation:

```{r}
gr = gr %>>% po("imputenewlvl")
```

and see what happens when we now train the whole **Graph**:

```{r}
task_imputed = gr$clone()$train(task)[[1]]
task_imputed$missings()
```

Awesome, no more missing values!

We could now use `task_imputed` for resampling and see whether a **ranger** model does better.
But this is dangerous! 
If we preprocess all training data at once, data could leak through the different cross-validation folds.
In order to do this properly, we have to process the training data in every fold separately.
Luckily, this is automatically handled in our `Graph`, if we use it through a `GraphLearner`.

We can simple append a `ranger` learner to the Graph and create a `GraphLearner` from this.

```{r}
glrn = GraphLearner$new(gr$clone() %>>% po(learner))
```

```{r}
res = resample(task, glrn, rdesc, store_models = TRUE)
agg = res$aggregate(msr("classif.acc"))
agg
```

So our model has not improved heavily, currently it has an accuracy of  `r round(agg, 3)`.
In order to improve more, we might need to do some feature engineering.

## Feature Engineering

We will do this using `PipeOpMutate` in order to showcase the power of `mlr3pipelines`.
Additionally, we will make use of the `character` columns, and thus re-select them:

```{r}
task$col_roles$feature = c(task$feature_names, c("cabin", "name", "ticket"))
```

```{r}
library("stringi")
po_ftextract = po("mutate", param_vals = list(
  mutation = list(
    fare_per_person = ~ fare / (parch + sib_sp + 1),
    deck = ~ factor(stri_sub(cabin, 1, 1)),
    title = ~ factor(stri_match(name, regex = ", (.*)\\.")[, 2]),
    surname = ~ factor(stri_match(name, regex = "(.*),")[, 2]),
    ticket_prefix = ~ factor(stri_replace_all_fixed(
      stri_trim(stri_match(ticket, regex = "(.*) ")[, 2]), ".", ""))
  )
))
```

Quickly checking what happens:

```{r}
task_eng = po_ftextract$clone()$train(list(task))[[1]]
task_eng$data()
```

Now we can put everything together again, we concatenate our new `PipeOp` with the `Graph` created abve
and use `PipeOpSelect` in order to de-select the `character` features we used for feature extraction.
Additionally, we collapse the 'surname', so only surnames that make up more than 0.6 \% of the data
are kept.

In summary we do the following: 

* `mutate`: The `po_ftextract` we defined above extracts additional features from the data.
* `collapsefactors`: Removes factor levels that make up less then 3 \% of the data.
* `select`: Drops `character` columns.
* `gunion`: Puts two `PipeOp`s in parallel.
  * `missind`: `po_indicator` adds a column for each numeric with the info whether the value is NA or not.
  * `imputehist`: Imputes numeric and integer columns by sampling from the histogram.
* `featureunion`: Cbind's parallel data streams.
* `imputenewlvl`: Imputes factor and ordered columns.
* `fixfactors`: Removes empty factor levels and removes factor levels that do not exist during training.
* `imputesample`: In some cases, if missing factor levels do not occur during training but only while predicting, `imputenewlvl` does not create a new level. For those, we sample a random value.
* `learner`: Appends a learner to the `Graph`.

The full graph we created is the following:

```{r}
learner = lrn("classif.ranger")
learner$param_set$values = list(num.trees = 500, min.node.size = 4)
```

```{r}
gr_final = po_ftextract %>>%
  po("collapsefactors", param_vals = list(no_collapse_above_prevalence = 0.03)) %>>%
  po("select", param_vals = list(selector = selector_invert(selector_type("character")))) %>>%
  gunion(list(po_indicator, po("imputehist"))) %>>%
  po("featureunion") %>>%
  po("imputenewlvl") %>>%
  po("fixfactors") %>>%
  po("imputesample") %>>%
  po(learner)
```

Let us see if things have improved:

```{r}
glrn = GraphLearner$new(gr_final)
res = resample(task, glrn, rdesc, store_models = TRUE)
agg = res$aggregate(msr("classif.acc"))
agg
```

We have improved even more! But there are many more things to explore!
We could extract even more information from the different features and see
what happens.

## Future

But now you are left to yourself! There are many [kaggle kernels](https://www.kaggle.com/c/titanic) that treat the **Titanic Dataset**
available. This can be a great starter to find even better models.
