---
title: A Pipeline for the Titanic Data Set - Basics
categories:
  - imputation
  - classification
  - mlr3pipelines
  - feature engineering
  - titanic data set
  - classification
author:
  - name: Florian Pfisterer
description: |
  This post shows how to build a Graph using the mlr3pipelines package on the "titanic" dataset.
date: 03-12-2020
output:
  distill::distill_article:
    self_contained: false
    highlight_downlit: true
    css: ../../custom.css
---

```{r 2020-03-12-mlr3pipelines-titanic-Basics-001, include=FALSE}
 knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 120)
)

library(mlr3book)
library(kableExtra)

ggplot2::theme_set(ggplot2::theme_light())
```

## Intro

We load the `r mlr_pkg("mlr3verse")` package which pulls in the most important packages for this example.
The `r mlr_pkg("mlr3learners")` package loads additional `r ref("Learner", "learners")`.
The data is part of the `r mlr_pkg("mlr3data")`  package.

```{r 2020-04-27-mlr3pipelines-titanic-advanced-002}
library(mlr3verse)
library(mlr3learners)
```

We initialize the random number generator with a fixed seed for reproducibility, and decrease the verbosity of the logger to keep the output clearly represented.

```{r 2020-04-27-mlr3pipelines-titanic-advanced-003}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
```

The titanic data is very interesting to analyze, even though it is part of many tutorials
and showcases.
This is because it requires many steps often required in real-world applications of machine
learning techniques, such as **missing value imputation**, **handling factors** and others.

Following features are illustrated in this use case section:

* Summarizing the data set
* Visualizing data
* Splitting data into train and test data sets
* Defining a task and a learner

## Exploratory Data Analysis

With the dataset, we get an explanation of the meanings of the different variables:

| Variables  | Description                         |
| ---------- | ----------------------------------- |
| `survived` | Survival                            |
| `name`     | Name                                |
| `age`      | Age                                 |
| `sex`      | Sex                                 |
| `sib_sp`   | Number of siblings / spouses aboard |
| `parch`    | Number of parents / children aboard |
| `fare`     | Amount paid for the ticket          |
| `pc_class` | Passenger class                     |
| `embarked` | Port of embarkation                 |
| `ticket`   | Ticket number                       |
| `cabin`    | Cabin                               |

We can use the `r cran_pkg("skimr")` package in order to get a first overview of the data:

```{r 2020-03-12-mlr3pipelines-titanic-Basics-004, R.options=list(width = 120)}
data("titanic", package = "mlr3data")

skimr::skim(titanic)
```

We can now create a `r ref("Task")` from our data.
As we want to classify whether the person survived or not, we will create a
`r ref("TaskClassif")`. We'll ignore the 'titanic_test' data for now and come back to it later.

## A first model

In order to obtain solutions comparable to official leaderboards, such as the ones available from kaggle, we split the data into train and validation set before doing any further analysis.
Here we are using the predefined split used by Kaggle.

```{r 2020-03-12-mlr3pipelines-titanic-Basics-007}
task = as_task_classif(titanic, target = "survived", positive = "yes")
task$set_row_roles(892:1309, "validation")
task
```

Our `r ref("Task")` currently has $3$ features of type `character`, which we don't really know how  to handle:
"Cabin", "Name", "Ticket" and "PassengerId".
Additionally, from our `r ref("skimr::skim()")` of the data, we have seen, that they have many unique values (up to 891).

We'll drop them for now and see how we can deal with them later on.

```{r 2020-03-12-mlr3pipelines-titanic-Basics-008}
task$select(cols = setdiff(task$feature_names, c("cabin", "name", "ticket")))
```

Additionally, we create a resampling instance that allows to compare data.

```{r 2020-03-12-mlr3pipelines-titanic-Basics-009}
cv3 = rsmp("cv", folds = 3L)$instantiate(task)
```

To get a first impression of what performance we can fit a simple decision tree:

```{r 2020-03-12-mlr3pipelines-titanic-Basics-010}
learner = mlr_learners$get("classif.rpart")
# or shorter:
learner = lrn("classif.rpart")

rr = resample(task, learner, cv3, store_models = TRUE)

rr$aggregate(msr("classif.acc"))
```

So our model should have a minimal accuracy of `0.80` in order to improve over the simple decision tree.
In order to improve more, we might need to do some feature engineering.

# Optimizing the model

If we now try to fit a 'ranger' random forest model, we will get an error,
as 'ranger' models can not naturally handle missing values.

```{r 2020-03-12-mlr3pipelines-titanic-Basics-011, error = TRUE}
learner = lrn("classif.ranger", num.trees = 250, min.node.size = 4)

rr = resample(task, learner, cv3, store_models = TRUE)
```

This means we have to find a way to impute the missing values.
To learn how to use more advanced commands of the mlr3pipelines package see:

* [Part II - Pipelines](https://mlr3gallery.mlr-org.com/posts/2020-04-27-mlr3pipelines-Imputation-titanic/)
