---
title: A pipeline for the titanic data set - Basics
categories:
  - imputation
  - classification
  - mlr3pipelines
  - feature engineering
author:
  - name: Florian Pfisterer
description: |
  This post shows how to build a Graph using the mlr3pipelines package on the "titanic" dataset.
date: 03-12-2020
output:
  distill::distill_article:
    self_contained: false
    css: ../../custom.css
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Just some preparation
knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 80)
)
if (require("data.table")) data.table::setDTthreads(1)
options(width = 90)
set.seed(8008135)
lgr::get_logger("mlr3")$set_threshold("warn")
library(skimr)
```

## Intro

First of all we are going to load required packages and the data. 
The data is part of the `mlr3data` package.

```{r}
library("mlr3")
library("mlr3learners")
library("mlr3pipelines")
library("mlr3data")
library("mlr3misc")
library("mlr3viz")
data("titanic")
```

The titanic data is very interesting to analyze, even though it is part of many tutorials 
and showcases. 
This is because it requires many steps often required in real-world applications of machine 
learning techniques, such as **missing value imputation**, **handling factors** and others.

Following features are illustrated in this use case section:
  
* Summarizing the data set
* Visualizing data
* Splitting data into train and test data sets
* Defining a task and a learner

In order to obtain solutions comparable to official leaderboards, such as the ones available from CRAN, we split the data into train and test set before doing any further analysis.
Here we are using the predefined split used by Kaggle. 

```{r}
titanic_train <- titanic[1:891, ]
titanic_test <- titanic[892:1309, ]
```

## Exploratory Data Analysis

With the dataset, we get an explanation of the meanings of the different variables:

```
survived        Survival
                (0 = No; 1 = Yes)
pclass          Passenger Class
                (1 = 1st; 2 = 2nd; 3 = 3rd)
name            Name
sex             Sex
age             Age
sibsp           Number of Siblings/Spouses Aboard
parch           Number of Parents/Children Aboard
ticket          Ticket Number
fare            Passenger Fare
cabin           Cabin
embarked        Port of Embarkation
                (C = Cherbourg; Q = Queenstown; S = Southampton)
```

We can use the `skimr` package in order to get a first overview of the data:

```{r, R.options=list(width = 120)}
skimr::skim(titanic_train)
skimr::skim(titanic_test)
```

Here we can also inspect the data for differences in the train and test set. 
This might be important, as shifts in the data distribution often make our models unreliable.

```{r, out.width="100%", fig.height=7}
DataExplorer::plot_bar(titanic_train, nrow = 5, ncol = 3)
```

```{r, out.width="100%", fig.height=4, warning=FALSE}
DataExplorer::plot_histogram(titanic_train, nrow = 2, ncol = 3)
DataExplorer::plot_boxplot(titanic_train, by = "survived", nrow = 2, ncol = 3)
```

We can now create a `Task` from our data. 
As we want to classify whether the person survived or not, we will create a 
`TaskClassif`. We'll ignore the 'titanic_test' data for now and come back to it later.

## A first model

```{r}
task <- TaskClassif$new("titanic", titanic_train, target = "survived", positive = "1")
task
```

Our `Task` currently has $3$ features of type `character`, which we don't really know how  to handle:
"Cabin", "Name", "Ticket" and "PassengerId".
Additionally, from our `skim` of the data, we have seen, that they have many unique values (up to 891).

We'll drop them for now and see how we can deal with them later on.

```{r}
task$select(cols = setdiff(task$feature_names, c("cabin", "name", "ticket")))
```

Additionally, we create a resampling instance that allows to compare data.

```{r}
rdesc <- rsmp("cv", folds = 3L)$instantiate(task)
```

To get a first impression of what performance we can fit a simple decision tree:

```{r}
learner <- mlr_learners$get("classif.rpart")
# or shorter:
learner <- lrn("classif.rpart")

res <- resample(task, learner, rdesc, store_models = TRUE)
agg <- res$aggregate(msr("classif.acc"))
agg
```

So our model should have a minimal accuracy of `r round(agg, 3)` in order to improve over the simple decision tree.
In order to improve more, we might need to do some feature engineering.

#Optimizing the model 

If we now try to fit a 'ranger' random forest model, we will get an error, 
as 'ranger' models can not naturally handle missing values.

```{r, error = TRUE}
learner <- lrn("classif.ranger")
learner$param_set$values <- list(num.trees = 250, min.node.size = 4)
res <- resample(task, learner, rdesc, store_models = TRUE)
```

This means we have to find a way to impute the missing values.
To learn how to use more advanced commands of the mlr3pipelines package see:
* [Part II - Pipelines](/2020-04-27-mlr3pipelines-titanic-advanced/)

<!-- Note that the link here needs to be fixed -->
