---
title: "Tuning a stacked learner"
categories:
  - mlr3pipelines
  - mlr3tuning
  - stacking
author:
  - name: Milan Dragicevic
  - name: Giuseppe Casalicchio
date: 04-27-2020
description: |
  This tutorial explains how to create and tune a multilevel stacking model using the mlr3pipelines package.
output:
  distill::distill_article:
    self_contained: false
    css: ../../custom.css
---

## Intro

```{r, echo = FALSE}
library(knitr)
opts_chunk$set(R.options = list(width = 100))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3filters)
library(mlr3tuning)
library(paradox)
library(glmnet)
```

Multilevel stacking is an ensemble technique, where predictions of several learners are added as new features to extend the orginal data on different levels.
On each level, the extended data is used to train a new level of learners.
This can be repeated for several iterations until a final learner is trained.
To avoid overfitting, it is advisable to use test set (out-of-bag) predictions in each level. 

In this post, a multilevel stacking example will be created using [mlr3pipelines](https://mlr3pipelines.mlr-org.com/articles/introduction.html) and tuned using [mlr3tuning](https://mlr3tuning.mlr-org.com/). 
A similar example is available in the [mlr3book](https://mlr3book.mlr-org.com/pipe-nonlinear.html#multilevel-stacking). 
However, we additionally explain how to tune the hyperparameters of the whole ensemble and each underlying learner jointly.  

In our stacking example, we proceed as follows:

1. **Level 0:** Based on the input data, we train three learners (`rpart`, `glmnet` and `lda`) on a sparser feature space obtained using different feature filter methods from [mlr3filters](https://mlr3filters.mlr-org.com/) to obtain slightly decorrelated predictions.
The test set predictions of these learners are attached to the original data (used in level 0) and will serve as input for the learners in level 1. 
2. **Level 1:** We transform this extended data using PCA, on which we then train additional three learners (`rpart`, `glmnet` and `lda`). 
The test set predictions of the level 1 learners are attached to input data used in level 1.
3. Finally, we train a final `ranger` learner to the data extended by level 1. Note that the number of features selected by the feature filter method in level 0 and the number of principal components retained in level 1 will be jointly tuned with some other hyperparameters of the learners in each level.

## Prerequisites

```{r packages}
library(mlr3)
library(mlr3learners)
library(mlr3pipelines)
library(mlr3filters)
library(mlr3tuning)
library(paradox)
library(glmnet)
```

For the stacking example, we use the [sonar classification task](https://mlr3.mlr-org.com/reference/mlr_tasks_sonar.html):

```{r task}
sonar_task = tsk("sonar")
sonar_task$col_roles$stratum = sonar_task$target_names #stratification
```

## Pipeline creation

### Level 0 

As mentioned, the level 0 learners are `rpart`, `glmnet` and `lda`:

```{r lrn1}
rprt_lrn  = lrn("classif.rpart", predict_type = "prob")
glmnet_lrn =  lrn("classif.glmnet", predict_type = "prob")
lda_lrn = lrn("classif.lda", predict_type = "prob")
```

To create the learner out-of-bag predictions, we use [PipeOpLearnerCV()](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_learner_cv.html):  

```{r lrn2}
rprt_cv1 = po("learner_cv", rprt_lrn, id = "rprt_1")
glmnet_cv1 = po("learner_cv", glmnet_lrn, id = "glmnet_1")
lda_cv1 = po("learner_cv", lda_lrn, id = "lda_1")
```

A sparser representation of the input data in level 0 is obtained using the following filters:  

```{r filt}
anova = po("filter", flt("anova"), id = "filt1")
mrmr = po("filter", flt("mrmr"), id = "filt2")
find_cor = po("filter", flt("find_correlation"), id = "filt3")
```

To summarize these steps into level 0, we use the [gunion](https://mlr3pipelines.mlr-org.com/reference/gunion.html) function. 
The out-of-bag predictions of all level 0 learners is attached using [PipeOpFeatureUnion](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_featureunion.html) along with the original data passed via [PipeOpNOP()](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_nop.html):  

```{r lrn3}
level0 = gunion(list(
  anova %>>% rprt_cv1,
  mrmr %>>% glmnet_cv1,
  find_cor %>>% lda_cv1,
  po("nop", id = "nop1")))  %>>%
  po("featureunion", id = "union1")
```

We can have a look at the graph from level 0:  

```{r lrn4, fig.width=6, fig.height = 6}
level0$plot(html = FALSE)
```

### Level 1

Now, we create the level 1 learners:

```{r lrn5}
rprt_cv2 = po("learner_cv", rprt_lrn , id = "rprt_2")
glmnet_cv2 = po("learner_cv", glmnet_lrn, id = "glmnet_2")
lda_cv2 = po("learner_cv", lda_lrn, id = "lda_2")
```

All level 1 learners will use [PCA](https://mlr3pipelines.mlr-org.com/reference/mlr_pipeops_pca.html) transformed data as input:

```{r lrn6}
level1 = level0 %>>%
  po("copy", 4) %>>%
  gunion(list(
    po("pca", id = "pca2_1", param_vals = list(scale. = TRUE)) %>>% rprt_cv2,
    po("pca", id = "pca2_2", param_vals = list(scale. = TRUE)) %>>% glmnet_cv2,
    po("pca", id = "pca2_3", param_vals = list(scale. = TRUE)) %>>% lda_cv2,
    po("nop", id = "nop2"))
  )  %>>%
  po("featureunion", id = "union2")
```

We can have a look at the graph from level 1:  

```{r lrn7, fig.width=6, fig.height = 7}
level1$plot(html = FALSE)
```

The out-of-bag predictions of the level 1 learners are attached to the input data from level 1 and a final ranger learner will be trained:  

```{r lrn8, fig.width=6, fig.height = 7}
ranger_lrn = lrn("classif.ranger", predict_type = "prob")

ensemble = level1 %>>% ranger_lrn
ensemble$plot(html = FALSE)
```

### Defining the tuning space

In order to tune the ensemble's hyperparameter jointly, we define the search space using [`ParamSet`](https://paradox.mlr-org.com/reference/ParamSet.html) from the [paradox](https://paradox.mlr-org.com/) package:

```{r ps1}
ps_ens = ParamSet$new(
  list(
    ParamInt$new("filt1.filter.nfeat", 5, 50),
    ParamInt$new("filt2.filter.nfeat", 5, 50),
    ParamInt$new("filt3.filter.nfeat", 5, 50),
    ParamInt$new("pca2_1.rank.", 3, 50),
    ParamInt$new("pca2_2.rank.", 3, 50),
    ParamInt$new("pca2_3.rank.", 3, 20),
    ParamDbl$new("rprt_1.cp", 0.001, 0.1),
    ParamInt$new("rprt_1.minbucket", 1, 10),
    ParamDbl$new("glmnet_1.alpha", 0, 1),
    ParamDbl$new("rprt_2.cp", 0.001, 0.1),
    ParamInt$new("rprt_2.minbucket", 1, 10),
    ParamDbl$new("glmnet_2.alpha", 0, 1),
    ParamInt$new("classif.ranger.mtry", lower = 1L, upper = 10L),
    ParamDbl$new("classif.ranger.sample.fraction", lower = 0.5, upper = 1),
    ParamInt$new("classif.ranger.num.trees", lower = 50L, upper = 200L)
  ))
```

### Performance comparison

Even with a simple ensemble, there is quite a few things to setup.
We compare the performance of the ensemble with a simple tuned ranger learner.

To proceed, we convert the `ensemble` pipeline as a [GraphLearner](https://mlr3pipelines.mlr-org.com/reference/mlr_learners_graph.html):  

```{r compare1}
ens_lrn = GraphLearner$new(ensemble)
ens_lrn$predict_type = "prob"
```

We define the search space for the simple ranger learner:

```{r compare2}
ps_ranger = ParamSet$new(
  list(
    ParamInt$new("mtry", lower = 1L, upper = 10L),
    ParamDbl$new("sample.fraction", lower = 0.5, upper = 1),
    ParamInt$new("num.trees", lower = 50L, upper = 200L)
  ))
```

For performance comparison, we use the [benchmark](https://mlr3.mlr-org.com/reference/benchmark.html) function that requires a design incorporating a list of learners and a list of tasks. 
Here, we have two learners (the simple ranger learner and the ensemble) and one task. 
Since we want to tune the simple ranger learner as well as the whole ensemble learner, we need to create an [AutoTuner](https://mlr3tuning.mlr-org.com/reference/AutoTuner.html) for each learner to be compared. 
To do so, we need to define a resampling strategy for the tuning in the inner loop (we use `cv3`) and for the final evaluation use use `outer_hold`:  

```{r compare3}
cv3 = rsmp("cv", folds = 3)

# AutoTuner for the ensemble learner
auto1 = AutoTuner$new(
    learner = ens_lrn,
    resampling = cv3,
    measures = msr("classif.auc"),
    tune_ps = ps_ens,
    terminator = term("evals", n_evals = 3), # to limit running time
    tuner = tnr("random_search")
  )

# AutoTuner for the simple ranger learner
auto2 = AutoTuner$new(
    learner = ranger_lrn,
    resampling = cv3,
    measures = msr("classif.auc"),
    tune_ps = ps_ranger,
    terminator = term("evals", n_evals = 3), # to limit running time
    tuner = tnr("random_search")
  )

# Define the list of learners
learns = list(auto1, auto2)

# For benchmarking, we use a simple holdout
set.seed(321)
outer_hold = rsmp("holdout")
outer_hold$instantiate(sonar_task)

design = benchmark_grid(
  tasks = sonar_task,
  learners = learns,
  resamplings = outer_hold
)

bmr = benchmark(design, store_models = TRUE)
bmr$aggregate(msr("classif.auc"))
```

For a more reliable comparison, the number of evaluation of the random search should be increased. 

## Conclusion

This example shows the versatility of [mlr3pipelines](https://mlr3pipelines.mlr-org.com/articles/introduction.html).
By using more learners, varied representations of the data set as well as more levels, a powerful yet compute hungry pipeline can be created. 
It is important to note that care should be taken to avoid name clashes of pipeline objects. 
