---
title: Tuning Over Multiple Learners
categories:
  - mlr3tuning
  - tuning
  - optimization
  - nested resampling
  - sonar credit data set
  - classification
author:
  - name: Jakob Richter
  - name: Bernd Bischl
date: 02-01-2020
description: |
  This use case shows how to tune over multiple learners for a single task.
output:
  distill::distill_article:
    self_contained: false
    highlight_downlit: true
    css: ../../custom.css
params:
  eval_all: FALSE
---

```{r 2020-02-01-tuning-multiplexer-001, include=FALSE}
 knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 120)
)

library(mlr3book)
library(kableExtra)

ggplot2::theme_set(ggplot2::theme_light())
```

This use case shows how to tune over multiple learners for a single task.
You will learn the following:

* Build a pipeline that can switch between multiple learners
* Define the hyperparameter search space for the pipeline
* Run a random or grid search (or any other tuner, always works the same)
* Run nested resampling for unbiased performance estimates

This is an advanced use case. What should you know before:

* `r mlr_pkg("mlr3")` basics
* `r mlr_pkg("mlr3tuning")` basics, especially `r ref("AutoTuner")`
* `r mlr_pkg("mlr3pipelines")`, especially branching

# The Setup

Assume, you are given some ML task and what to compare a couple of learners, probably because you want to select the best of them at the end of the analysis.
That's a super standard scenario, it actually sounds so common that you might wonder: Why an (advanced) blog post about this? With pipelines?
We will consider 2 cases: (a) Running the learners in their default, so without tuning, and (b) with tuning.

We load the `r mlr_pkg("mlr3verse")` package which pulls in the most important packages for this example.
The `r mlr_pkg("mlr3learners")` package loads additional `r ref("Learner", "learners")`.

```{r 2020-02-01-tuning-multiplexer-002}
library(mlr3verse)
library(mlr3learners)
```

We initialize the random number generator with a fixed seed for reproducibility, and decrease the verbosity of the logger to keep the output clearly represented.

```{r 2020-02-01-tuning-multiplexer-003}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

Let's define our learners.

```{r 2020-02-01-tuning-multiplexer-004}
learners = list(
  lrn("classif.xgboost", id = "xgb", eval_metric = "logloss"),
  lrn("classif.ranger", id = "rf")
)
learners_ids = sapply(learners, function(x) x$id)

task = tsk("sonar") # some random data for this demo
inner_cv2 = rsmp("cv", folds = 2) # inner loop for nested CV
outer_cv5 = rsmp("cv", folds = 5) # outer loop for nested CV
```

# Default Parameters

# The Benchmark-Table Approach
Assume we don't want to perform tuning and or with running all learner in their respective defaults.
Simply run benchmark on the learners and the tasks. That tabulates our results nicely and shows us what works best.

```{r 2020-02-01-tuning-multiplexer-005}
grid = benchmark_grid(task, learners, outer_cv5)
bmr = benchmark(grid)
bmr$aggregate(measures = msr("classif.ce"))
```

# The Pipelines Approach

Ok, why would we ever want to change the simple approach above - and use pipelines / tuning for this?
Three reasons:

1. What we are doing with `r ref("benchmark()")` is actually statistically flawed, insofar if we report the error of the numerically best method from the benchmark table as its estimated future performance.
   If we do that we have "optimized on the CV" (we basically ran a grid search over our learners!) and we know that this is will produce optimistically biased results.
   NB: This is a somewhat ridiculous criticism if we are going over only a handful of options, and the bias will be very small.
   But it will be noticeable if we do this over hundreds of learners, so it is important to understand the underlying problem.
   This is a somewhat subtle point, and this gallery post is more about technical hints for `r mlr_pkg("mlr3")`, so we will stop this discussion here.
2. For some tuning algorithms, you might have a chance to more efficiently select from the set of algorithms than running the full benchmark.
   Because of the categorical nature of the problem, you will not be able to learn stuff like "If learner A works bad, I don't have to try learner B", but you can potentially save some resampling iterations.
   Assume you have so select from 100 candidates, experiments are expensive, and you use a 20-fold CV.
   If learner A has super-bad results in the first 5 folds of the CV, you might already want to stop here.
   "Racing" would be such a tuning algorithm.
3. It helps us to foreshadow what comes later in this post where we tune the learners.


The pipeline just has a single purpose in this example:
It should allow us to switch between different learners, depending on a hyperparameter.
The pipe consists of three elements:

* `r ref("mlr_pipeops_branch", text = "branch")` pipes incoming data to one of the following elements, on different data channels. We can name these channel on construction with `options`.
* our learners (combined with `r ref("gunion()")`)
* `r ref("mlr_pipeops_unbranch", text = "unbranch")` combines the forked paths at the end.

```{r 2020-02-01-tuning-multiplexer-006, fig.width = 10, fig.height = 8}
graph =
  po("branch", options = learners_ids) %>>%
  gunion(lapply(learners, po)) %>>%
  po("unbranch")
graph$plot()
```

The pipeline has now quite a lot of available hyperparameters.
It includes all hyperparameters from all contained learners.
But as we don't tune them here (yet), we don't care (yet).
But the first hyperparameter is special.
`branch.selection` controls over which (named) branching channel our data flows.

```{r 2020-02-01-tuning-multiplexer-007}
graph$param_set$ids()
graph$param_set$params$branch.selection
```

We can now tune over this pipeline, and probably running grid search seems a good idea to "touch" every
available learner.
NB: We have now written down in (much more complicated code) what we did before with `benchmark`.

```{r 2020-02-01-tuning-multiplexer-008, results = "hide"}
graph_learner = as_learner(graph)
graph_learner$id = "g"
graph_learner$param_set$values$branch.selection = to_tune(levels = c("rf", "xgb"))

instance = tune(
  method = "grid_search",
  task = task,
  learner = graph_learner,
  resampling = inner_cv2,
  measure = msr("classif.ce"))

as.data.table(instance$archive)
```

```{r 2020-02-01-tuning-multiplexer-009, echo = FALSE}
as.data.table(instance$archive, exclude_columns = c("uhash",  "resample_result")) %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE)
```

But: Via this approach we can now get unbiased performance results via nested resampling and using the `r ref("AutoTuner")` (which would make much more sense if we would select from 100 models and not 2).

```{r 2020-02-01-tuning-multiplexer-010, eval = params$eval_all}
at = auto_tuner(
  method = "grid_search",
  learner = graph_learner,
  resampling = inner_cv2,
  measure = msr("classif.ce"),
)

rr = resample(task, at, outer_cv5, store_models = TRUE)

# access inner tuning result
extract_inner_tuning_results(rr)
```

```{r 2020-02-01-tuning-multiplexer-011, echo = FALSE, eval = params$eval_all}
saveRDS(rr, "data/rr.rda")
```

```{r 2020-02-01-tuning-multiplexer-012, echo = FALSE}
rr = readRDS("data/rr.rda")
```

```{r 2020-02-01-tuning-multiplexer-013, echo = FALSE}
extract_inner_tuning_results(rr)[, .(iteration, branch.selection, classif.ce, task_id, learner_id, resampling_id )] %>%
  kable(format = "html") %>%
  kable_styling()
```

```{r 2020-02-01-tuning-multiplexer-014, results = 'hide'}
# access inner tuning archives
extract_inner_tuning_archives(rr)
```

```{r 2020-02-01-tuning-multiplexer-015, echo = FALSE, layout="l-page"}
extract_inner_tuning_archives(rr, exclude_columns = c("uhash",  "resample_result")) %>%
  kable(format = "html") %>%
  kable_styling()
```

# Model-Selection and Tuning with Pipelines

Now let's select from our given set of models and tune their hyperparameters.
One way to do this is to define a search space for each individual learner, wrap them all with the `r ref("AutoTuner")`, then call `r ref("benchmark()")` on them.
As this is pretty standard, we will skip this here, and show an even neater option, where you can tune over models and hyperparameters in one go.
If you have quite a large space of potential learners and combine this with an efficient tuning algorithm, this can save quite some time in tuning as you can learn during optimization which options work best and focus on them.
NB: Many AutoML systems work in a very similar way.

# Define the Search Space

Remember, that the pipeline contains a joint set of all contained hyperparameters.
Prefixed with the respective PipeOp ID, to make names unique.

```{r 2020-02-01-tuning-multiplexer-016, eval = FALSE}
as.data.table(graph$param_set)
```

```{r 2020-02-01-tuning-multiplexer-017, echo = FALSE}
as.data.table(graph$param_set)[, .(id, class, lower, upper, nlevels)] %>%
  kable(format = "html") %>%
  kable_styling(full_width = TRUE) %>%
  scroll_box(height = "400px", extra_css = "border: 0px !important;")
```

We decide to tune the `mtry` parameter of the random forest and the `nrounds` parameter of xgboost.
Additionally, we tune branching parameter that selects our learner.

We also have to reflect the hierarchical order of the parameter sets (admittedly, this is somewhat inconvenient).
We can only set the `mtry` value if the pipe is configured to use the random forest (`r ref("mlr_learners_classif.ranger", "ranger")`).
The same applies for the xgboost parameter.

```{r 2020-02-01-tuning-multiplexer-018}
search_space = ps(
  branch.selection = p_fct(c("rf", "xgb")),
  rf.mtry = p_int(1L, 20L, depends = branch.selection == "rf"),
  xgb.nrounds = p_int(1, 500, depends = branch.selection == "xgb"))
```

# Tune the Pipeline with a Random Search

Very similar code as before, we just swap out the search space.
And now use random search.

```{r 2020-02-01-tuning-multiplexer-019, eval = params$eval_all}
graph_learner = as_learner(graph)
graph_learner$id = "g"

instance = tune(
  method = "random_search",
  task = task,
  learner = graph_learner,
  resampling = inner_cv2,
  measure = msr("classif.ce"),
  search_space = search_space,
  term_evals = 10
)

as.data.table(instance$archive)
```

```{r 2020-02-01-tuning-multiplexer-020, echo = FALSE, eval = params$eval_all}
saveRDS(instance, "data/instance.rda")
```

```{r 2020-02-01-tuning-multiplexer-021, echo = FALSE}
instance = readRDS("data/instance.rda")
```

```{r 2020-02-01-tuning-multiplexer-022, echo = FALSE, layout="l-page"}
as.data.table(instance$archive, exclude_columns = c("uhash",  "resample_result")) %>%
  kable(format = "html") %>%
  kable_styling()
```

The following shows a quick way to visualize the tuning results.

```{r 2020-02-01-tuning-multiplexer-023, fig.width=10, fig.height=5}
autoplot(instance, cols_x = c("xgb.nrounds","rf.mtry"))
```

Nested resampling, now really needed:

```{r 2020-02-01-tuning-multiplexer-024, eval = params$eval_all}
rr = tune_nested(
  method = "grid_search",
  task = task,
  learner = graph_learner,
  inner_resampling = inner_cv2,
  outer_resampling = outer_cv5,
  measure = msr("classif.ce"),
  search_space = search_space,
  term_evals = 10L)
```

```{r 2020-02-01-tuning-multiplexer-025, echo = FALSE, eval = params$eval_all}
saveRDS(rr, "data/rr_2.rda")
```

```{r 2020-02-01-tuning-multiplexer-026, echo = FALSE}
rr = readRDS("data/rr_2.rda")
```

```{r 2020-02-01-tuning-multiplexer-027, eval = FALSE}
# access inner tuning result
extract_inner_tuning_results(rr)
```

```{r 2020-02-01-tuning-multiplexer-028, echo = FALSE}
extract_inner_tuning_results(rr) %>%
  kable(format = "html") %>%
  kable_styling()
```
