---
title: mlr3 basics - german credit
author: Martin Binder, Florian Pfisterer & Philipp Kopper
date: '2020-03-11'
slug: basics_german_credit
categories: []
tags: ['visualization', 'random forest', 'classification', 'feature importance']
packages: ['mlr3', 'mlr3learners', 'skimr', 'DataExplorer']
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Just some preparation
knitr::opts_chunk$set(
  cache = FALSE,
  collapse = TRUE,
  comment = "#>"
)
if (require("data.table")) data.table::setDTthreads(1)
options(width = 90)
set.seed(8008135)
lgr::get_logger("mlr3")$set_threshold("warn")
library(skimr)
```

## Intro

This is the first part in a serial of tutorials.
The other parts of this series can be found here:

* [Part II - Pipelines](/basics_pipelines_german_credit/)
* [Part III - Tuning](/basics_tuning_german_credit/)

We will walk through this tutorial interactively. The text is kept short to be followed in real time. 

## Prerequisites

Ensure all packages used in this tutorial are installed. This includes packages from the `mlr3` family, as well as other tools for data handling, cleaning and visualisation which we are going to use. [Appendix for more info about packages](#packages)

<!--
```{r, message=FALSE, warning=FALSE, eval = FALSE}
# install from CRAN
packages_cran = c("remotes", "data.table", "ggplot2", "skimr", "DataExplorer",
  "precrec", "farff", "curl", "visNetwork", "kknn", "MASS", "ranger",
  "xgboost", "e1071", "future", "future.apply")

# install things from GitHub that are not yet on CRAN
packages_gith = "mlr-org/mlr3viz"

to_install = setdiff(packages_cran, installed.packages()[,"Package"])
if (length(to_install)) install.packages(to_install)
install.packages(c("mlr3", "mlr3misc", "paradox", "mlr3filters", "mlr3learners",
  "mlr3pipelines", "mlr3tuning"))
lapply(packages_gith, remotes::install_github)
```
-->

Load the packages we are going to use:

```{r, message=FALSE, warning=FALSE}
library("data.table")
library("mlr3")
library("mlr3learners")
library("ggplot2")
```

```{r, include=FALSE}
theme_set(theme_light())
```

## Machine Learning Use Case: German Credit Data

In our tutorials we use the `german_credit` dataset for our next example.
The German credit data is a research data set of the University of Hamburg from 1994 donated by Prof. Hans Hoffman.
The data features personal, demographic and financial data on individuals.
Furthermore, it contains an assessment on the individual's creditworthniness.
A detailed description (and manual download) can be found at the [UCI repository](https://archive.ics.uci.edu/ml/datasets/Statlog+%28German+Credit+Data%29).

As the `"GermanCredit"` data set is frequently used within `r mlr_pkg("mlr3")`, there is a task already existing just for this purpose.

- Goal is to classify people by their credit risk (**good** or **bad**) using 20 features:

| Feature Name             | Description                                            |
| ------------------------ | ------------------------------------------------------ |
| `checking_status`        | status/balance of checking account at this bank        |
| `duration`               | duration of the credit in months                       |
| `credit_history`         | past credit history of applicant at this bank          |
| `purpose`                | reason customer is applying for a loan                 |
| `credit_amount`          | amount asked by applicant                              |
| `savings_status`         | savings accounts/bonds at this bank                    |
| `employment`             | present employment since                               |
| `installment_commitment` | installment rate in percentage of disposable income    |
| `personal_status`        | combination of sex and personal status of applicant    |
| `other_parties`          | other debtors/guarantors present?                      |
| `residence_since`        | present residence since                                |
| `property_magnitude`     | properties that applicant has                          |
| `age`                    | age in years                                           |
| `other_payment_plans`    | other installment plans the applicant is paying        |
| `housing`                | type of apartment rented, owned, for free / no payment |
| `existing_credits`       | number of existing credits at this bank                |
| `job`                    | current job information                                |
| `num_dependents`         | number of people being liable to provide maintenance   |
| `own_telephone`          | is there any telephone registered for this customer?   |
| `foreign_worker`         | is applicant foreign worker?                           |

### Importing the Data

The dataset we are going to look at is the **german credit** dataset. It comes with the **mlr3** package.
We can load it using the `tsk()` function.

```{r, message=FALSE}
task = tsk("german_credit")
```

Before we use the data, we'll do a thorough investigation of the dataset.

```{r, message=FALSE}
credit = task$data()
```

### Exploring the Data
We can get a quick overview on our data set using R's summary function:

```{r 09-use-cases-2-002}
german_credit = task$data()
summary(german_credit)
dim(german_credit)
```

Our dataset has `r nrow(german_credit)` observations and `r ncol(german_credit)` columns.
The variable we want to predict is `credit_risk` (either **good** or **bad**).
That means we aim to classify people by their credit risk (**good** or **bad**).
  
We recommend the `skimr` (`skimr::skim`) and `DataExplorer` (`DataExplorer::plot_bar`, `DataExplorer::plot_histogram`, `DataExplorer::plot_boxplot`) packages as they create very well readable and understandable overviews.

```{r, R.options=list(width = 120)}
skimr::skim(credit)
```

```{r, echo = FALSE}
library(mlr3misc)
credit_short = map_dtc(credit, function(x) {
  if (is.factor(x) || is.character(x)) {
    levels(x) = abbreviate(stringr::str_trunc(levels(x), 16), 12)
  }
  x
})
```

<!-- Visible code --> 
```{r, eval = FALSE, out.width="100%", fig.height=7}
p = DataExplorer::plot_bar(credit_short, nrow = 5, ncol = 3)
```

<!-- Invisible code -->
```{r, echo = FALSE, out.width="100%", fig.height=7}
p = DataExplorer::plot_bar(credit_short, nrow = 5, ncol = 3)
```

```{r, out.width="100%", fig.height=4}
DataExplorer::plot_histogram(credit, nrow = 2, ncol = 3)
DataExplorer::plot_boxplot(credit, by = "credit_risk", nrow = 2, ncol = 3)
```

During this exploratory analysis meaningful analysis aspects could be:

  - Skewed distributions
  - Missing values
  - Empty / rare factor variables
  
An explanatory analysis is crucial to get a feeling for your data.
On the other hand the data can be validated this way.
Non plausible data can be investigated or outliers can be removed.
  
However, in this use case we do not focus on the exploratory analysis.

We want to do modeling now!

## Modeling
Considering how we are going to tackle the problem relates closely to what `mlr3` entities we will use.

The typical questions that arise when building an ML workflow are:

What is the problem we are trying to solve?
What are appropriate learning algorithms?
How do we evaluate "good" performance? 

More systematically in `r mlr_pkg("mlr3")` they can be expressed via five components:

1) The **task** defintion

2) The **learner** definition

3) The **training**

4) The **prediction**

5) The evaluation via **measures**

### Task Definition

First of all, we are interested in the target which we want to model.
Most supervised machine learning problems are __regression__ or __classification__ problems.
However, note that other problems include unsupervised learning or time-to-event data (covered in `mlr3survival`).
In the `r mlr_pkg("mlr3")` context to distinguish between these problems, we define **tasks**.
If we want to solve a classification problem, we define a classifaction task -- `TaskClassif`.
For a regression problem, we establish a regression task -- `TaskRegr`.
Tasks are not only a different word for problems.
They also reflect some technical properties with respect to `r mlr_pkg("mlr3")`.

In our case it is clearly our objective to model or predict the binary `factor` variable `credit_risk`.
Thus, we use `TaskClassif`.

With `TaskClassif$new()` we can initialize a classification task. 
We need to specify a name for the task (we choose `"GermanCredit"`), the data we want to use (`credit`) and name of the target (`"credit_risk"`):

```{r}
task = TaskClassif$new("GermanCredit", credit, "credit_risk")
```

### Learner Definition

After having decided _what_ should be modeled, we need to decide on how.
This means we need to decide which learning algrorithms, or **Learners** are appropriate.
Using prior knowledge (e.g. knowing that it is a classification task or assuming that the classes are linearly separable) one ends up with one or some suitable **Learners**.
Many learners can be obtained via the `mlr3learners` package (CRAN). 
Additionally, many learners are provided on the `mlr3learners` organsisation [github page](https://github.com/mlr3learners).
These two ressources combined account for a large fraction of standard learning algorithms.
As `r mlr_pkg("mlr3")` usually only wraps learners from packages, it is easy to create a formal **Learner** by yourself using our template (). 
You may find the section 'Extending mlr3' in this book also very helpful.
If you happen to write your own **Learner** in `r mlr_pkg("mlr3")`, we would be happy if you share it with the `r mlr_pkg("mlr3")` community.

All available `Learner`s (i.e. all which you have installed from `r mlr_pkg("mlr3")`, `mlr3learners`, the mlr3learners github page or self-written ones) can be obtained by:

```{r}
mlr_learners
```

For our problem, a suitable learner could be one of the following:
Logistic regression, CART, Random Forest (or many more).

A learner can be initialized with the `lrn` function and the name of the learner (e.g., `lrn("classif.xxx")`). 
Use `?mlr_learners_xxx` to open the help page of a learner named `xxx`, or use [the internet](https://mlr3learners.mlr-org.com/reference/index.html).

For example a logistic regression can be initialised in the following manner (logistic regression using R's `glm()` function and is provided by the `mlr3learners` package).

```{r}
learner_logreg = lrn("classif.log_reg")
```

Alternatively but equivalently, one could use:

```{r}
learner_logreg = LearnerClassifLogReg$new()
```

### Training

Training is the procedure, where a model is fitted on data.
The selected model is trained so that is reflects the data situation in the best possible manner.

#### Logistic Regression

We start with the example of the logistic regression.
However, you will immediately see that the procedure generalises to any learner very easily.

An initialised learner can be trained on data using the `R6` method `train`.

```{r}
learner_logreg$train(task)
```

Typically, in machine learning, one does not use the full data which is available but a subset, the so-called training data.

To efficiently perform a spilt of the data one could do the following:

```{r}
train_set = sample(task$nrow, 0.8 * task$nrow)
test_set = setdiff(seq_len(task$nrow), train_set)
```

80 percent of the data is used for training.
The remaining 20 percent are used for evaluation at a subsequent later point in time.
`train_set` is an integer vector displaying the selected rows of the original data set.

```{r}
head(train_set)
```

In `r mlr_pkg("mlr3")` the training with a subset of the data is declared by the additional argument `row_ids = train_set`.

```{r}
learner_logreg$train(task, row_ids = train_set)
```

We used logistic regression from the `r cran_pkg("stats")` package.
To be precise `stats::glm` with the `family` argument set to `family = binomial()`.
All other arguments are set to the default of the `r ref("stats::glm")` function.
`r mlr_pkg("mlr3")` only works as a wrapper. 
The original model can be accessed via:

```{r}
learner_logreg$model
```

The stored object is a normal `glm` object and works with its `S3` methods.

```{r}
class(learner_logreg$model)
summary(learner_logreg$model)
```

#### Random Forest

Just like the logistic regression, we could train a random forest instead.
We use the fast implementation from the `r cran_pkg("ranger")` package.
To do so we first need to define the learner and then actually train it.
Before we just used the default logistic regression learner.

We now additionally supply the importance argument (`importance = "permutation"`).
Doing so, we override the default and let the learner do feature importance determination based on permutation feature importance.

```{r}
learner_rf = lrn("classif.ranger", importance = "permutation")
learner_rf$train(task)
```

We can access the importance values using `$importance()`.
The `R6` importance method plots the feature importance derived during training.

```{r}
learner_rf$importance()
```

In order to obtain a better plot, we convert the importance into a `data.table`.

```{r}
importance = as.data.table(learner_rf$importance(), keep.rownames = TRUE)
colnames(importance) = c("Feature", "Importance")
importance
ggplot(importance, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col() + coord_flip() + xlab("")
```

### Prediction

After training the machine learning model, the model can be used for prediction.
Usually, prediction is the main purpose of machine learning models.
In our case, the model can be used to classify new credit applicants w.r.t. their associated credit status (good vs. bad) on the basis of the features.
Typically, machine learning models create numeric values.
In the regression case this is very natural.
For classification, most models create scores or probabilites.
Based on these values, one can derive class predictions.

#### Predict Classes

First, we directly predict classes.
Here, we do something which is almost never a good idea in machine learning:
Use the same data for prediction as for training.
However, for the purpose of demonstration we can negelct this.

```{r}
pred_logreg = learner_logreg$predict_newdata(german_credit)
pred_rf = learner_rf$predict_newdata(german_credit)
pred_logreg
pred_rf
```

The `predict()` function gives a `Prediction` object. 
It can be converted to a `data.table` if one wants to use it downstream.

We display the prediction results aggregated in a confusion matrix.

```{r}
pred_logreg$confusion
pred_rf$confusion
```

The random forest seems to perform much better than the logistic regression.
Based on this output, we would always want the random forest to do our credit worthiness classification.
However, as outlined, we evalaute on the training data.
It is possible that the results are steered by overfitting.
Additionally, we use the default options of the packages which seems 'unfair'.
It could be that the default options of the random forest are much better suited to the data situation than the logistic regression.
We should do performance evaluation, tuning and benchmarks

#### Predict Probabilities

Most learners may not only predict a class variable ("response"), but also their degree of "belief" / uncertainty in a given response.
Typically, we achieve this by setting the `$predict_type` slot to `"prob"`.
Sometimes this needs to be done *before* the learner is trained.
This is necessary because the wrapped package requires this.
Alternatively, we can directly create the learner with this option: `lrn("classif.log_reg", predict_type = "prob")`.

```{r}
learner_logreg$predict_type = "prob"
```

```{r}
learner_logreg$predict_newdata(german_credit)
```

Note that sometimes one needs to be cautious when dealing with the probability interpretation of the predictions.

### Performance Evaluation

To measure the performance of a learner on new unseen data, we usually mimic the scenario of unseen data by splitting up the data into training and test set.
The training set is used for training the learner, and the test set is only used for predicting and evaluating the performance of the trained learner.
Numerous resampling methods (cross-validation, bootstrap) repeat the splitting process in different ways.

We need to specify the resampling strategy using the `rsmp()` function:

```{r}
resampling = rsmp("holdout", ratio = 2 / 3)
print(resampling)
```

Here, we use "holdout", a simple train-test split (with just one interation).
We use the `resample()` function to undertake the resampling calculation:

```{r}
res = resample(task, learner_logreg, resampling)
res
```

The default score of the measure is included in the `$aggregate()` slot:

```{r}
res$aggregate()
```

The default in this scenario is `classif.ce`. 
This refers to the classification error. 
A low value close to one is desired.

We can easily run differend resampling strategies, e.g. repeated holdout (`"subsampling"`), or cross validation.
Most methods perform repeated train/predict cycles on different data subsets and aggregate the result (usually as the `mean()`). 
Doing this manually would require us to write loops.
`r mlr_pkg("mlr3")` does the job for us.

```{r}
res_sub = resample(task, learner_logreg, rsmp("subsampling", repeats = 10))
res_sub$aggregate()
```

Instead, we could also apply cross-validation.

```{r}
res_cv = resample(task, learner_logreg, rsmp("cv", folds = 10))
res_cv$aggregate()
```

`r mlr_pkg("mlr3")` features scores for many more measures. 
Here we apply `msr("classif.fpr")` for the false positive rate, or `msr("classif.fnr")` for the false negative rate.
Multiple measures are entered via a list.

```{r}
# false positive rate
res_cv$aggregate(msr("classif.fpr"))
# false positive rate and false negative
msr_list = list(
  msr("classif.fpr"),
  msr("classif.fnr")
)
res_cv$aggregate(msr_list)
```

There are a few more resampling methods, and quite a few more measures. 
We list them in

```{r}
mlr_resamplings
```

and 

```{r}
mlr_measures
```

To get help on a resampling method, use `?mlr_resamplings_xxx`, for a measure do `?mlr_measures_xxx`. 
You can also use the [mlr3 reference](https://mlr3.mlr-org.com/reference/index.html) online.

Some measure, for example `"auc"`, require a "probability" prediction, instead of a response prediction.

### Performance Comparison and Benchmarks

We could compare `Learners` by evaluating `resample()` for each of them manually.
However, `benchmark()` automatically performs resampling evaluations for multiple learners and tasks.
`benchmark_grid()` creates fully crossed designs:
Multiple `Learner`s for multiple `Task`s are compared w.r.t. multiple `Resampling`s.
```{r}
lrn_list = list(
  lrn("classif.log_reg", predict_type = "prob"),
  lrn("classif.ranger", predict_type = "prob")
)
bm_design = benchmark_grid(task = task, resamplings = rsmp("cv", folds = 10), learners = lrn_list)
```

At this point, we want to mention that large benchmarks will (naturally) be very time-consuming and computionally expensvie.
Our use case is very simple, so expect it to take about one minute.
In general, we want use *parallelization* to speed things up on multicore machines.
A benchmark is an 'embarassingly parallel' task, so in this case it would be very easy.
This does **not work** on rstudio cloud, though.
So only uncomment it locally!

```{r}
# future::plan("multiprocess")
```

```{r}
bmr = benchmark(bm_design)
```

In the benchmark we can compare different measures. 
We compare misclassification rate and AUC.

```{r}
msr_list = list(msr("classif.ce"), msr("classif.auc"))
performances = bmr$aggregate(msr_list)
performances[, c("learner_id", "classif.ce", "classif.auc")]
```

In contrast to before, we see that the two models perform very similarly.
The perfect class assignment from before was obviously a very optimistic evalution.

## Deviating from package defaults

The previously shown techniques build the backbone of an `r mlr_pkg("mlr3")`-featured machine learning work bench.
However, in most cases one would never proceed in the way we did.
While many `R` packages have carefully selected default settings, they will not perform optimally in any scenario.
Typically, we can select the values of hyperparameters.
These (hyper)parameters can be accessed via

```{r}
learner_logreg$param_set$get_values()
```

or 

```{r}
learner_logreg$param_set
```

As of now, the parameter set is empty.

We can parametrise our learners in two distinct manners.
If we have prior knowledge on how the learner should be (hyper-)parametrised, the way to go would be manually entering the parameters in the parameter set.
In most cases, however, we would want to tune the model so that it can search good model configurations itself.
For now however we only want to compare a small number of models.

To get an idea on which parameters can be manipulated, we can investigate the parameters of the original package version.

```{r}
?ranger::ranger
```

```{r}
expression1 = floor(sqrt(ncol(german_credit) - 1))
``` 

For the random forest two meaninful parameters which steer model complexity are:
`num.trees` and `mtry`

`num.trees` defaults to `500` and `mtry` to `floor(sqrt(ncol(data) - 1))`, in our case `r expression1`.

In the following we aim to train three different learners:

1) The default random forest.

2) A random forest with low `num.trees` and low `mtry`.

3) A random forest with high `num.trees` and high `mtry`.

We will benchmark their performance on the German credit data set.

We construct the three learners and set the parameters with a named list.

```{r}
learner_rf_medium = lrn("classif.ranger", predict_type = "prob")

learner_rf_low = lrn("classif.ranger", predict_type = "prob")
learner_rf_low$param_set$values <- list(num.trees = 5, mtry = 2)

learner_rf_high = lrn("classif.ranger", predict_type = "prob")
learner_rf_high$param_set$values <- list(num.trees = 1000, mtry = 11)
```

Once the learners are defined, we can benchmark them.

```{r}
lrn_list = list(
  learner_rf_low = learner_rf_low,
  learner_rf_medium = learner_rf_medium,
  learner_rf_high = learner_rf_high
)
bm_design = benchmark_grid(task = task, resamplings = rsmp("cv", folds = 10), learners = lrn_list)
```

```{r}
bmr = benchmark(bm_design)
```

We compare misclassification rate and AUC again.

```{r}
msr_list = list(msr("classif.ce"), msr("classif.auc"))
performances = bmr$aggregate(msr_list)
cbind(learner = names(lrn_list), performances[, c("classif.ce", "classif.auc")])
```

All learners seem to have similar performance while the `learner_rf_low` seems to underfit a bit.

## Outlook

This tutorial was very detailed on the default workflow within `r mlr_pkg("mlr3")`.
Having followed this tutorial you should be able to run your first models yourself.
Next to that we spiked into performance evaluation and benchmarking.
Furthermore, we showed how to customise learners.

The next parts of the tutorial will go more into depth into additional `mlr3` topics.

[Part II - Pipelines](/basics_pipelines_german_credit/) introduces you into the `r mlr_pkg("mlr3pipelines")` package.
You will learn you to build complete machine learning workflows including, e.g. data preprocessing.

[Part III - Tuning](/basics_tuning_german_credit/) introduces you into the `r mlr_pkg("mlr3tuning")` package.
You will learn how to find parametrisations of your model that perform better than others.

## Appendix

### R Pro Tips

* What are the arguments of `lrn()`, `tsk()`, etc. again? -> Think about the corresponding dictionary.

```{r}
mlr_learners

mlr_tasks

mlr_measures

mlr_resamplings
```

* What are the arguments of a `$new()` constructor?

```{r}
formals(TaskClassif$public_methods$initialize)
```

* What are the possible slots and functions of an object?

```{r}
# Writing `pred_rf$`, and pressing <TAB> should work.
# Otherwise:
names(pred_rf)

# try names without `()` first
# and see if it is a function
```

* How do I see the help file of an object

```{r}
# The documentation is organized by object classes
class(pred_rf)

# use ?PredictionClassif, ?Prediction etc.
# Try all elements listed in the class
```

### mlr3 and its Extensions

| Package         | Functionality                                                                     |
| :-------------- | :-------------------------------------------------------------------------------- |
| `mlr3`          | Framework for machine learning: `Task`, `Learner`, `resample()` and `benchmark()` |
| `mlr3learners`  | Concrete `Learner`s for many popular machine learning implementations             |
| `mlr3pipelines` | Dataflow programming of machine learning workflows.                               |
| `mlr3tuning`    | Hyperparameter tuning for machine learning algorithms.                            |
| `mlr3filter`    | Feature filtering                                                                 |
| `mlr3viz`       | Visualisations and plots                                                          |
| `paradox`       | Auxiliary package providing (hyper)parameter handling                             |
| `mlr3misc`      | Auxiliary functions        

### Packages

The non-`mlr3` packages we use:

| Package         | Reason                                                                                                                                                                   |
| :-------------- | :----------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| `remotes`       | We use this only to be able to do `remotes::install_github()`. This enables us to install packages from GitHub that are not on CRAN yet.                                 |
| `data.table`    | This provides a more efficient and versatile replacement for the `data.frame` datatype built into R. [Intro vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) |
| `ggplot2`       | A very powerful plotting tool. [Overview with link to "cheat sheets"](https://ggplot2.tidyverse.org/)                                                                    |
| `callr`         | Encapsulating function calls in external R sessions. [GitHub page](https://github.com/r-lib/callr#readme)                                                                |
| `future`        | Parallelization to make use of multicore functionality. [GitHub page](https://github.com/HenrikBengtsson/future)                                                         |
| `skimr`         | Plotting data summaries for exploratory data analysis. [Vignette](https://cran.r-project.org/web/packages/skimr/vignettes/Using_skimr.html)                              |
| `DataExplorer`  | Plotting data for exploratory data analysis. [Vignette](https://cran.r-project.org/web/packages/DataExplorer/vignettes/dataexplorer-intro.html)                          |
| `rpart.plot`    | Plotting CART trees. [Website](http://www.milbo.org/rpart-plot/)                                                                                                         |
| `precrec`       | Plotting AUC curves. [Vignette](https://cran.r-project.org/web/packages/precrec/vignettes/introduction.html)                                                             |
| `glmnet`        | Provides the `"*.glmnet"` `Learner`s. Penalized regression is often surprisingly powerful, especially in high-dimensional settings.                                      |
| `kknn`          | Provides the `"*.kknn"` `Learner`s. k-nearest neighbor classification / regression is a classical machine learning technique.                                            |
| `MASS`          | Provides the `"*.lda"` and `"*.qda"` `Learner`s.                                                                                                                         |
| `ranger`        | Provides the `"*.ranger"` `Learner`s. This is an implementation of the powerful "Random Forest" algorithm, which often works very well, even without parameter tuning.   |
| `xgboost`       | Provides the `"*.xgboost"` `Learner`s. Gradient boosting is often among the best performing machine learning methods, although it may require parameter tuning.          |
| `e1071`         | Provides the `"*.svm"` and `"classif.naive_bayes"` `Learner`s. SVMs (support vector machines) perform well, but are very dependent on correctly chosen kernel parameters.|
