---
title: Feature Selection on the Titanic Data Set
categories:
  - mlr3fselect
  - optimization
  - feature selection
  - nested resampling
  - titanic data set
  - classification
description: |
  We give a short introduction to mlr3fselect.
author:
  - name: Marc Becker
date: 01-08-2021
output:
  distill::distill_article:
    self_contained: false
    css: ../../custom.css
params:
  eval_all: FALSE
---

```{r practical-tuning-series-tune-a-support-vector-machine-001, include=FALSE}
 knitr::opts_chunk$set(
  echo = TRUE,
  R.options = list(width = 120)
)

library(mlr3book)
library(kableExtra)

set.seed(7832)

ggplot2::theme_set(ggplot2::theme_light())
```

```{r mlr3fselect-basic-002, include=FALSE, eval = params$eval_all}
future::plan("multicore")
```

## Introduction

In this tutorial, we introduce the `r mlr_pkg("mlr3fselect")` package by comparing feature selection methods on the Titanic disaster data set.
The objective of feature selection is to enhance the interpretability of models, speed up the learning process and increase the predictive performance.

We load the `r mlr_pkg("mlr3verse")` package which pulls in the most important packages for this example.

```{r mlr3fselect-basic-003}
library(mlr3verse)
```

We initialize the random number generator with a fixed seed for reproducibility, and decrease the verbosity of the logger to keep the output clearly represented.

```{r mlr3fselect-basic-004}
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
lgr::get_logger("bbotk")$set_threshold("warn")
```

## Titanic Data Set

The [Titanic data set](https://www.kaggle.com/c/titanic/data) contains data for 887 Titanic passengers, including whether they survived when the Titanic sank.
Our goal will be to predict the survival of the Titanic passengers.

After loading the data set from the `r mlr_pkg("mlr3data")` package, we impute the missing age values with the median age of the passengers, set missing embarked values to `"s"` and remove `character` features.
We could use feature engineering to create new features from the `character` features, however we want to focus on feature selection in this tutorial.

In addition to the `survived` column, the reduced data set contains the following attributes for each passenger:

| Feature    | Description                         |
|------------|-------------------------------------|
| `age`      | Age                                 |
| `sex`      | Sex                                 |
| `sib_sp`   | Number of siblings / spouses aboard |
| `parch`    | Number of parents / children aboard |
| `fare`     | Amount paid for the ticket          |
| `pc_class` | Passenger class                     |
| `embarked` | Port of embarkation                 |

```{r mlr3fselect-basic-005}
library(mlr3data)

data("titanic", package = "mlr3data")
titanic$age[is.na(titanic$age)] = median(titanic$age, na.rm = TRUE)
titanic$embarked[is.na(titanic$embarked)] = "S"
titanic$ticket = NULL
titanic$name = NULL
titanic$cabin = NULL
titanic = titanic[!is.na(titanic$survived),]
```

We construct a binary classification task.

```{r mlr3fselect-basic-006}
task = as_task_classif(titanic, target = "survived", positive = "yes")
```

## Model

We use the logistic regression learner provided by the `r mlr_pkg("mlr3learners")` package.

```{r mlr3fselect-basic-007}
library(mlr3learners)

learner = lrn("classif.log_reg")
```

To evaluate the predictive performance, we choose a 3-fold cross-validation and the classification error as the measure.

```{r mlr3fselect-basic-008}
resampling = rsmp("cv", folds = 3)
measure = msr("classif.ce")

resampling$instantiate(task)
```

## Classes

The `r ref("FSelectInstanceSingleCrit")` class specifies a general feature selection scenario.
It includes the `r ref("ObjectiveFSelect")` object that encodes the black box objective function which is optimized by a feature selection algorithm.
The evaluated feature sets are stored in an `r ref("ArchiveFSelect")` object.
The archive provides a method for querying the best performing feature set.

The `r ref("Terminator")` classes determine when to stop the feature selection.
In this example we choose a terminator that stops the feature selection after 10 seconds.
The sugar functions `r ref("trm()")` and `r ref("trms()")` can be used to retrieve terminators from the `r ref("mlr_terminators")` dictionary.

```{r mlr3fselect-basic-009}
terminator = trm("run_time", secs = 10)
FSelectInstanceSingleCrit$new(
  task = task,
  learner = learner,
  resampling = resampling,
  measure = measure,
  terminator = terminator)
```

The `r ref("FSelector")` subclasses describe the feature selection strategy. The sugar function `r ref("fs()")` can be used to retrieve feature selection algorithms from the `r ref("mlr_fselectors")` dictionary.

```{r mlr3fselect-basic-010}
mlr_fselectors
```

## Random search

Random search randomly draws feature sets and evaluates them in batches.
We retrieve the `r ref("FSelectorRandomSearch")` class with the `r ref("fs()")` sugar function and choose `r ref("TerminatorEvals")`.
We set the `n_evals` parameter to `10` which means that 10 feature sets are evaluated.

```{r mlr3fselect-basic-011}
terminator = trm("evals", n_evals = 10)
instance = FSelectInstanceSingleCrit$new(
  task = task,
  learner = learner,
  resampling = resampling,
  measure = measure,
  terminator = terminator)
fselector = fs("random_search", batch_size = 5)
```

The feature selection is started by passing the `r ref("FSelectInstanceSingleCrit")` object to the `$optimize()` method of `r ref("FSelectorRandomSearch")` which generates the feature sets.
These features set are internally passed to the `$eval_batch()` method of `r ref("FSelectInstanceSingleCrit")` which evaluates them with the objective function and stores the results in the archive.
This general interaction between the objects of **mlr3fselect** stays the same for the different feature selection methods.
However, the way how new feature sets are generated differs depending on the chosen `r ref("FSelector")` subclass.

```{r mlr3fselect-basic-012}
fselector$optimize(instance)
```

The `r ref("ArchiveFSelect")` stores a `data.table::data.table()` which consists of the evaluated feature sets and the corresponding estimated predictive performances.

```{r mlr3fselect-basic-013}
as.data.table(instance$archive, exclude_columns = c("runtime_learners", "timestamp", "batch_nr", "resample_result", "uhash"))
```

The associated resampling iterations can be accessed in the `r ref("BenchmarkResult")` by calling

```{r mlr3fselect-basic-014}
instance$archive$benchmark_result
```

We retrieve the best performing feature set with

```{r mlr3fselect-basic-015}
instance$result
```

## Sequential forward selection

We try sequential forward selection. We chose `r ref("TerminatorStagnation")` that stops the feature selection if the predictive performance does not increase anymore.

```{r mlr3fselect-basic-016}
terminator = trm("stagnation", iters = 5)
instance = FSelectInstanceSingleCrit$new(
  task = task,
  learner = learner,
  resampling = resampling,
  measure = measure,
  terminator = terminator)

fselector = fs("sequential")
fselector$optimize(instance)
```

The `r ref("FSelectorSequential")` object has a special method for displaying the optimization path of the sequential feature selection.

```{r mlr3fselect-basic-017}
fselector$optimization_path(instance)
```

## Recursive feature elimination

Recursive feature elimination utilizes the `$importance()` method of learners.
In each iteration the feature(s) with the lowest importance score is dropped.
We choose the non-recursive algorithm (`recursive = FALSE`) which calculates the feature importance once on the complete feature set.
The recursive version (`recursive = TRUE`) recomputes the feature importance on the reduced feature set in every iteration.

```{r mlr3fselect-basic-018}
learner = lrn("classif.ranger", importance = "impurity")
terminator = trm("none")
instance = FSelectInstanceSingleCrit$new(
  task = task,
  learner = learner,
  resampling = resampling,
  measure = measure,
  terminator = terminator,
  store_models = TRUE)

fselector = fs("rfe", recursive = FALSE)
fselector$optimize(instance)
```

We access the results.

```{r mlr3fselect-basic-019, eval = FALSE}
as.data.table(instance$archive, exclude_columns = c("runtime_learners", "timestamp", "batch_nr", "resample_result", "uhash"))
```

```{r mlr3fselect-basic-020, echo = FALSE}
as.data.table(instance$archive, exclude_columns = c("runtime_learners", "timestamp", "batch_nr", "resample_result", "uhash")) %>%
  kable(format = "html") %>%
  kable_styling() %>%
  kableExtra::scroll_box(width = "100%", height = "200px")
```

## Nested resampling

It is a common mistake to report the predictive performance estimated on resampling sets during the feature selection as the performance that can be expected from the combined feature selection and model training.
The repeated evaluation of the model might leak information about the test sets into the model and thus leads to over-fitting and over-optimistic performance results.
[Nested resampling](https://mlr3book.mlr-org.com/nested-resampling.html) uses an outer and inner resampling to separate the feature selection from the performance estimation of the model.
We can use the `r ref("AutoFSelector")` class for running nested resampling.
The `r ref("AutoFSelector")` essentially combines a given `r ref("Learner")` and feature selection method into a `r ref("Learner")` with internal automatic feature selection.
The inner resampling loop that is used to determine the best feature set is conducted internally each time the `r ref("AutoFSelector")` `r ref("Learner")` object is trained.

```{r mlr3fselect-basic-021}
resampling_inner = rsmp("cv", folds = 5)
measure = msr("classif.ce")

at = AutoFSelector$new(
  learner = learner,
  resampling = resampling_inner,
  measure = measure,
  terminator = terminator,
  fselect = fs("sequential"),
  store_models = TRUE)
```

We put the `r ref("AutoFSelector")` into a `r ref("resample()")` call to get the outer resampling loop.

```{r mlr3fselect-basic-022, eval = params$eval_all}
resampling_outer = rsmp("cv", folds = 3)

rr = resample(task, at, resampling_outer, store_models = TRUE)
```

```{r mlr3fselect-basic-023, echo = FALSE, eval = params$eval_all}
saveRDS(rr, "data/rr.rda")
```

```{r mlr3fselect-basic-024, echo = FALSE}
rr = readRDS("data/rr.rda")
```

The aggregated performance of all outer resampling iterations is the unbiased predictive performance we can expected from the logistic regression model with an optimized feature set found by sequential selection.

```{r mlr3fselect-basic-025}
rr$aggregate()
```

We check whether the feature sets that were selected in the inner resampling are stable.
The selected feature sets should not differ too much.
We might observe unstable models in this example because the small data set and the low number of resampling iterations might introduces too much randomness.
Usually, we aim for the selection of similar feature sets for all outer training sets.

```{r mlr3fselect-basic-026, eval = FALSE}
extract_inner_fselect_results(rr)
```

```{r mlr3fselect-basic-027, echo = FALSE, layout="l-body-outset"}
extract_inner_fselect_results(rr) %>%
  kable(format = "html") %>%
  kable_styling() %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

Next, we want to compare the predictive performances estimated on the outer resampling to the inner resampling.
Significantly lower predictive performances on the outer resampling indicate that the models with the optimized feature sets overfit the data.

```{r mlr3fselect-basic-028}
rr$score()[, .(iteration, task_id, learner_id, resampling_id, classif.ce)]
```

The archives of the `r ref("AutoFSelector")`s gives us all evaluated hyperparameter configurations (i.e. feature sets) with the associated predictive performances.

```{r mlr3fselect-basic-029, eval = FALSE}
extract_inner_fselect_archives(rr)
```

```{r mlr3fselect-basic-030, echo = FALSE, layout="l-page"}
extract_inner_fselect_archives(rr) %>%
  kable(format = "html") %>%
  kable_styling() %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

## Shortcuts

Selecting a feature subset can be shortened by using the `r ref("fselect()")`-shortcut.

```{r mlr3fselect-basic-031, eval = FALSE}
instance = fselect(
  method = "random_search",
  task = tsk("iris"),
  learner = lrn("classif.log_reg"),
  resampling = rsmp("cv", folds = 3),
  measure = msr("classif.ce"),
  term_evals = 10
)
```

Applying nested resampling can be shortened by using the `r ref("fselect_nested()")`-shortcut.

```{r mlr3fselect-basic-032, eval = FALSE}
rr = fselect_nested(
  method = "random_search",
  task = tsk("iris"),
  learner = lrn("classif.log_reg"),
  inner_resampling = rsmp ("cv", folds = 3),
  outer_resampling = rsmp("cv", folds = 3),
  measure = msr("classif.ce"),
  term_evals = 10
)
```
