---
title: "Feature Selection on the Titanic Data Set"
categories:
  - mlr3fselect
  - optimization
  - feature selection
  - nested resampling
description: |
  We give a short introduction to mlr3fselect.
author:
  - name: Marc Becker
date: 01-08-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (require("data.table")) data.table::setDTthreads(1)
set.seed(7832)
lgr::get_logger("mlr3")$set_threshold("warn")
```

## Introduction

In this tutorial, we introduce the **mlr3fselect** package by comparing feature selection methods on the Titanic disaster data set. 
The objective of feature selection is to enhance the interpretability of models, speed up the learning process and increase the predictive performance.

## Titanic data set

The [Titanic data set](https://www.kaggle.com/c/titanic/data) contains data for 887 Titanic passengers, including whether they survived when the Titanic sank. 
Our goal will be to predict the survival of the titanic passengers. 

After loading the data set from the `mlr3data` package, we impute the missing age values with the median age of the passengers, set missing embarked values to `s` and remove `character` features.
We could use feature engineering to create new features from the `character` features, however we want to focus on feature selection in this tutorial.

In addition to the `survived` column, the reduced data set contains for each passenger the following attributes:

* `age` Age 
* `sex` Sex 
* `sib_sp` Number of siblings/ spouses aboard
* `parch` Number of parents/ children aboard
* `fare` Fair paid 
* `pc_class` Passenger class
* `embarked` Port of embarkation

```{r}
library(mlr3data)

data("titanic", package = "mlr3data")
titanic$age[is.na(titanic$age)] = median(titanic$age, na.rm = TRUE)
titanic$embarked[is.na(titanic$embarked)] = "S"
titanic$ticket = NULL
titanic$name = NULL
titanic$cabin = NULL
titanic = titanic[!is.na(titanic$survived),]
```

We construct a binary classification task.

```{r}
library(mlr3)

task = TaskClassif$new(id = "titanic", backend = titanic, target = "survived", positive = "yes")
```

## Model

We use the logistic regression learner provided by the `mlr3learners` package.

```{r}
library(mlr3learners)

learner = lrn("classif.log_reg")
```

To evaluate the predictive performance, we choose a `3-fold-cross-validation` and the classification error as the measure.

```{r}
resampling = rsmp("cv", folds = 3)
measure = msr("classif.ce")

resampling$instantiate(task)
```

## Classes

The `FSelectInstanceSingleCrit` class specifies a general feature selection scenario. 
It includes the  `ObjectiveFSelect` object that encodes the black box objective function which is optimized by a feature selection algorithm. 
The evaluated feature sets are stored in an `ArchiveFSelect` object. 
The archive provides a method for querying the best performing feature set.

The `Terminator` subclasses determine when to stop the feature selection. 
In this example we choose a terminator that stops the feature selection after 10 seconds. 
The sugar functions `trm()` and `trms()` can be used to retrieve terminators from the `mlr_terminators` dictionary.

```{r}
library(mlr3fselect)

terminator = trm("run_time", secs = 10)
FSelectInstanceSingleCrit$new(
  task = task, 
  learner = learner, 
  resampling = resampling, 
  measure = measure, 
  terminator = terminator)
```

The `FSelector` subclasses describe the feature selection strategy. The sugar function `fs()` can be used to retrieve feature selection algorithms from the `mlr_fselectors` dictionary.

```{r}
mlr_fselectors
```

## Random search

Random search randomly draws feature sets and evaluates them in batches. 
We retrieve the `FSelectRandomSearch` class with the `fs()` sugar function and choose the `evals` terminator. 
We set the `n_evals` parameter to `10` which means that 10 feature sets are evaluated.

```{r}
terminator = trm("evals", n_evals = 10)
instance = FSelectInstanceSingleCrit$new(
  task = task, 
  learner = learner, 
  resampling = resampling, 
  measure = measure, 
  terminator = terminator)
fselector = fs("random_search", batch_size = 5)
```

The feature selection is started by passing the `FSelectInstanceSingleCrit` object to the `$optimize()` method of `FSelectorRandomSearch` which generates the feature sets. 
These features set are internally passed to the `$eval_batch()` method of `FSelectInstanceSingleCrit` which evaluates them with the objective fuction and stores the results in the archive. 
This general interaction between the objects of **mlr3fselect** stays the same for the different feature selection methods.
However, the way how new feature sets are generated differs depending on the chosen `FSelector` subclass.

```{r}
fselector$optimize(instance)
```

The `ArchiveFSelect` stores a `data.table` which conists of the evaluated feature sets and the corresponding estimated predictive performances. 

```{r}
as.data.table(instance$archive)[, 1:8]
```

The associated resampling iterations can be accessed in the `BenchmarkResult` by calling

```{r}
instance$archive$benchmark_result
```

We retrieve the best performing feature set with 

```{r}
instance$result
```

## Sequential forward selection 

We try sequential forward selection. We chose the `stagnation` terminator that stops the feature selection if the predictive performance does not increase anymore.

```{r}
terminator = trm("stagnation", iters = 5)
instance = FSelectInstanceSingleCrit$new(
  task = task, 
  learner = learner, 
  resampling = resampling, 
  measure = measure, 
  terminator = terminator)

fselector = fs("sequential")
fselector$optimize(instance)
```

The `FSelectorSequential` object has a special method for displaying the optimization path of the sequential feature selection. 

```{r}
fselector$optimization_path(instance)
```

## Recursive feature elimination

Recrusive feature elemiation utilizes the `$importance()` method of learners.
In each iteration the feature(s) with the lowest importance score is droped. 
We choose the non-recruvie algorithm (`recursive = FALSE`) which calculates the feature importance once on the complete feature set. 
The recrusive version (`recursive = TRUE`) recomputes the feature importance on the reduced feature set in every iteration. 

```{r}
learner = lrn("classif.ranger", importance = "impurity")
terminator = trm("none")
instance = FSelectInstanceSingleCrit$new(
  task = task, 
  learner = learner, 
  resampling = resampling, 
  measure = measure, 
  terminator = terminator,
  store_models = TRUE)

fselector = fs("rfe", recursive = FALSE)
fselector$optimize(instance)
```

We access the results.

```{r}
as.data.table(instance$archive)[, 1:8]
```

## Nested resampling

It is a common mistake to report the predictive performance estimated on resampling sets during the feature selection as the performance that can be expected from the model.
The repeated evaluation of the model might leak information about the test sets into the model and thus leads to overfitting and over-optimistic performance results. 
[Nested resampling](https://mlr3book.mlr-org.com/nested-resampling.html) uses an outer and inner resampling to separate the feature selection from the performance estimation of the model.
We can use the `AutoFselector` class for running nested resampling.
The `AutoFselector` wraps a `Learner` and argument it with an automatic feature selection.
This is our inner resampling loop.

```{r}
resampling_inner = rsmp("cv", folds = 5)
measure = msr("classif.ce")

at = AutoFSelector$new(
  learner = learner, 
  resampling = resampling_inner, 
  measure = measure,
  terminator = terminator,
  fselect = fs("sequential"),
  store_models = TRUE)
```

We put the `AutoFSelector` into a `resample()` call to get the outer resampling loop.

```{r}
resampling_outer = rsmp("cv", folds = 3)

rr = resample(task, at, resampling_outer, store_models = TRUE)
```

The aggregated performance of all outer resampling iterations is the unbiased predictive performance we can expected from the logisitic regression model with an optimized feature set found by sequential selection. 

```{r}
rr$aggregate()
```

Next, we want to compare the predictive performances estimated on the outer resampling to the inner resampling.
Significantly lower predictive performances on the outer resampling indicate that the models with the optimized feature sets overfit the data.

```{r}
rr$score()
```

We check if the feature sets are stable that were selected in the inner resampling. 
The selected featute sets should not differ too much.
We might observe unstable models in this example because the small data set and the low number of resampling iterations might introduces too much randomness. 
Usually, we aim for the selection of similiar feature sets for all outer training sets.

```{r}
do.call(rbind, lapply(rr$learners, function(x) x$fselect_result))
```

The archive of the `AutoFSelector` gives us all evaluated hyperparameter configurations with the associated predictive performances.

```{r}
rr$learners[[1]]$archive$data[, 1:8]
```
